---
title: "R code for rhythm ratios analyses and plots for Chimpanzee buttress drumming shows rhythmicity and subspecies variation"
author: Jelle van der Werff and Vesta Eleuteri
date: 31/08/2024
output:
    html_notebook:
        code_folding: hide
        toc: true
        toc_float: true
---

# Preliminaries

```{r load-packages}
rm(list=ls())

# for easy loading of packages
if (!require("pacman")) install.packages("pacman")

# load packages
p_load("here", "ggplot2", "dplyr", "nlme", "kldtools", "pracma")
library(conflicted) 

# for relative file paths (e.g. here('Input', 'file.csv'))
here::i_am(file.path("Drum_ratios_rcode.Rmd"))
conflicts_prefer(here::here) 

# load data
ratios <- read.csv(here("dataframes", "drumming_ratios.csv"))

# load custom ggplot theme
source(here("Drumming_ggplot_theme.R"))

# create simulated data
ratios.simulated <- runif(n = 100001, min = 100, max = 1000)

# calculate ratios
ratios.simulated <- ratios.simulated / (ratios.simulated + lead(ratios.simulated))
ratios.simulated <- ratios.simulated[!is.na(ratios.simulated)]
ratios.simulated <- as.data.frame(list(Ratio = ratios.simulated))
```

# Coding ratios categories

```{r define-on-off}
labels <- as.character(seq(0, 5))
breaks <- c(0,
1 / 2.5, 1 / 2.25, 1 / 2, (1 - 1 / 2.25), (1 - 1 / 2.5),
  1
)
ratios$Bin <- cut(ratios$Ratio, breaks = breaks, labels = labels)
ratios.simulated$Bin <- cut(ratios.simulated$Ratio, breaks = breaks, labels = labels)

# "0" Other
# "1"  OFF 1:1
# "2"  "3"  ON 1:1
# "4"  OFF 1:1
# "5" Other
```

# Counting on- vs. off-isochronous

```{r count-and-normalize}
# Code off- and on-isochronous
ratios$On_off_isoc <- NA
ratios$On_off_isoc[ratios$Bin %in% c(1, 4)] <- "Off"
ratios$On_off_isoc[ratios$Bin %in% c(2, 3)] <- "On"
ratios$On_off_isoc[ratios$Bin %in% c(0, 5)] <- "Other"
# Calculate normalized counts
ratios.isoc.counts <- dplyr::summarize(dplyr::group_by(ratios, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)
# Normalize
ratios.isoc.counts$On_isochronous_norm <- ratios.isoc.counts$On_isochronous / (ratios.isoc.counts$Total_obs * (0.0444 * 2))
ratios.isoc.counts$Off_isochronous_norm <- ratios.isoc.counts$Off_isochronous / (ratios.isoc.counts$Total_obs * (0.0555 * 2))
```

#Find Local Maxima 

```{r local maxima}
# Number of points at which density is estimated
n_steps <- 1000

# Get the ratios separate for both subspecies
ratios.eac <- ratios$Ratio[ratios$Subspecies == "EAC"]
ratios.wac <- ratios$Ratio[ratios$Subspecies == "WAC"]

# Get the density of the ratios
dens.eac <- density(ratios.eac, from = 0, to = 1, n = n_steps)
dens.wac <- density(ratios.wac, from = 0, to = 1, n = n_steps)

# Find local maxima
maxima.eac <- findpeaks(dens.eac$y, nups = 1, ndowns = 1)
maxima.wac <- findpeaks(dens.wac$y, nups = 1, ndowns = 1)

# Get the x values
maxima.x.eac <- maxima.eac[, 2] / n_steps
maxima.x.wac <- maxima.wac[, 2] / n_steps
```
# Plotting

## EAC vs. WAC
```{r plot-ratios}
# select only on-isochronous and off-isochronous
# ratios <- ratios[ratios$Bin %in% c(1, 2, 3, 4), ]
# ratios.simulated <- ratios.simulated[ratios.simulated$Bin %in% c(1, 2, 3, 4), ]

# tiles
tiles <- as.data.frame(list(
  xmin = c(0, breaks),
  xmax = c(breaks, 1),
  ymin = rep(-Inf, length(breaks) + 1),
  ymax = rep(Inf, length(breaks) + 1),
  fill = as.factor(c("Other", "Off 1:1", "Off 1:1", "On 1:1", "On 1:1", "Off 1:1", "Off 1:1", "Other"))
))

labels <- as.data.frame(list(
  x = c(0.42, 0.5, 0.58),
  y = c(0.1, 0.1, 0.1),
  label = c("Off 1:1", "On 1:1", "Off 1:1")
))


str(ratios)

# rename Subspecies for plot
ratios$Subspecies=as.factor(ratios$Subspecies)
ratios$Subspecies2=as.factor(ratios$Subspecies)
levels(ratios$Subspecies2)[levels(ratios$Subspecies2)=="WAC"] <- "Western" #relevel
levels(ratios$Subspecies2)[levels(ratios$Subspecies2)=="EAC"] <- "Eastern" #relevel

str(ratios)

# plot subspecies
sub_ratios_plot=ggplot(ratios, aes(x = Ratio, fill = factor(Subspecies2, levels=c("Eastern", "Western")))) +
  geom_density(data = subset(ratios, Subspecies2 == "Western"), 
  aes(x = Ratio, fill = "Western"), 
  linewidth = 0, alpha = 0.7, position = "identity") +
   geom_density(data = subset(ratios, Subspecies2 == "Eastern"), 
  aes(x = Ratio, fill = "Eastern"), 
  linewidth = 0, alpha = 0.7, position = "identity") +
  geom_density(data = ratios.simulated, aes(x = Ratio), linewidth = 2, colour = "yellow", inherit.aes = FALSE, show.legend = FALSE) +
  theme_drumming() +
    scico::scale_colour_scico_d(palette = "roma", begin=0.1, end=0.9, alpha=0.7) + # fourth item in list
    scico::scale_fill_scico_d(palette = "roma", begin=0.1, end=0.9, alpha=0.7) +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.8, colour = "dark grey") +
  geom_rect(data = tiles, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill="grey", inherit.aes = FALSE, alpha = 0.2,  show.legend = FALSE) + 
  labs(y = "Density", x="r", fill="", title = "A - Subspecies rhythmic patterns") + 
  geom_label(data = labels, aes(x = x, y = y, label = label), size = 10, family = "Helvetica", inherit.aes = FALSE) + theme( 
    plot.title = element_text(size = 38),  
    axis.title.x = element_text(size = 32), 
    axis.title.y = element_text(size = 32),  
    axis.text.x = element_text(size = 28),  
    axis.text.y = element_text(size = 28), 
    legend.title = element_text(size = 32),
    legend.text = element_text(size = 32)  
  ) + theme(plot.margin = margin(1, 0, 1, 1, "cm"),
  legend.position = c(0.05, 0.99), # Position legend in the top left corner
  legend.justification = c(0, 1),  # Align the legend to the top left corner
  legend.background = element_blank(),  # Remove legend background
  legend.box.background = element_blank() # Remove legend box background
  ) 

ggsave(here("plots", "main plots", "Sub_integer_ratios_vs_random2.png"), width = 26, height = 10)
ggsave(here("plots", "main plots", "Sub_integer_ratios_vs_random2.svg"), width = 26, height = 10)
```

# Kolmogorov-Smirnov Tests for rhythmic drumming

## Bonferroni correction for multiple testing

```{r bonferroni-cor}
alpha <- 0.05  
num_comparisons <- 3
alpha_corrected <- alpha / num_comparisons #0.01666667
```

## Is chimpanzee drumming random?

```{r drum-ks-test}

set.seed(1)
ks.drum <- ksboot(ratios$Ratio, ratios.simulated$Ratio, alternative = "two.sided")
ks.drum 

# Save list of results
Chimps_KS_res <- list(
  p_value = ks.drum$ksboot.pvalue,
  bootstraps = ks.drum$nboots)

# Convert the list to a data frame
Chimps_KS_res <- as.data.frame(Chimps_KS_res)

# Save the data frame to a CSV file
write.table(Chimps_KS_res, file = "All_KS_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")
```
KS-test is significant, i.e. the drumming isn´t random. 

## Is the drumming random within and across subspecies?

### KS: EAC

```{r eac-ks-test}
set.seed(2)
ks.eac <- ksboot(ratios$Ratio[ratios$Subspecies == "EAC"], ratios.simulated$Ratio, alternative = "two.sided")
ks.eac 

# Save list of results
EAC_KS_res <- list(
  p_value = ks.eac$ksboot.pvalue,
  bootstraps = ks.eac$nboots)

# Convert the list to a data frame
EAC_KS_res <- as.data.frame(EAC_KS_res)

# Save the data frame to a CSV file
write.table(EAC_KS_res, file = "EAC_KS_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")
```

KS-test is significant, i.e. the drumming isn´t random. 

### KS: WAC

```{r wac-ks-test}
set.seed(3)
ks.wac <- ksboot(ratios$Ratio[ratios$Subspecies == "WAC"], ratios.simulated$Ratio, alternative = "two.sided")
ks.wac

# Save list of results
WAC_KS_res <- list(
  p_value = ks.wac$ksboot.pvalue,
  bootstraps = ks.wac$nboots)

# Convert the list to a data frame
WAC_KS_res <- as.data.frame(WAC_KS_res)

# Save the data frame to a CSV file
write.table(WAC_KS_res, file = "WAC_KS_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")
```

KS-test is significant, i.e. the drumming isn´t random.


### KS: EAC vs. WAC

But are the two distributions different? 

```{r eac-wac-ks-test}
ks.eac.vs.wac <- ksboot(ratios$Ratio[ratios$Subspecies == "EAC"], ratios$Ratio[ratios$Subspecies == "WAC"], alternative = "two.sided")
ks.eac.vs.wac

# Save list of results
EACvsWAC_KS_res <- list(
  p_value = ks.eac.vs.wac$ksboot.pvalue,
  bootstraps = ks.eac.vs.wac$nboots)

# Convert the list to a data frame
EACvsWAC_KS_res <- as.data.frame(EACvsWAC_KS_res)

# Save the data frame to a CSV file
write.table(EACvsWAC_KS_res, file = "EACvsWAC_KS_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")
```
Yes

# Wilcoxon Tests for isochronous drumming

## Bonferroni correction for multiple testing

```{r bonferroni-cor}
alpha <- 0.05  
num_comparisons <- 3
alpha_corrected <- alpha / num_comparisons #0.01666667
```

### More on-isochronous vs. off-isochronous drumming across and within subspecies?

```{r isoc-counts-wilcoxon}
# Two-sample Wilcoxon test

# All chimpanzees
wx.drum=wilcox.test(ratios.isoc.counts$On_isochronous_norm, ratios.isoc.counts$Off_isochronous_norm, paired = TRUE)

  ##Save list of results
Chimps_WX_res <- list(
  statistic = wx.drum$statistic, 
  p_value = wx.drum$p.value)
                      
  ##Convert the list to a data frame
Chimps_WX_res <- as.data.frame(Chimps_WX_res)

  ##Save the data frame to a CSV file
write.table(Chimps_WX_res, file = "All_WX_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")
median(ratios.isoc.counts$On_isochronous_norm)
median(ratios.isoc.counts$Off_isochronous_norm)

# East African chimpanzees
wx.eac=wilcox.test(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"], ratios.isoc.counts$Off_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"], paired = TRUE)

  ##Save list of results
EAC_WX_res <- list(
  statistic = wx.eac$statistic, 
  p_value = wx.eac$p.value)
                      
  ##Convert the list to a data frame
EAC_WX_res <- as.data.frame(EAC_WX_res)

  ##Save the data frame to a CSV file
write.table(EAC_WX_res, file = "EAC_WX_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")

median(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"])
median(ratios.isoc.counts$Off_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"])

# West African chimpanzees
wx.wac=wilcox.test(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"], ratios.isoc.counts$Off_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"], paired = TRUE)

  ##Save list of results
WAC_WX_res <- list(
  statistic = wx.wac$statistic, 
  p_value = wx.wac$p.value)
                      
  ##Convert the list to a data frame
WAC_WX_res <- as.data.frame(WAC_WX_res)

  ##Save the data frame to a CSV file
write.table(WAC_WX_res, file = "WAC_WX_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")

median(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"])
median(ratios.isoc.counts$Off_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"])
```

### More isochrony in WAC than EAC?

```{r isoc-counts-wilcoxon}
# Two-sample Wilcoxon test
wx.eac.vs.wac=wilcox.test(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"], ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"])

  ##Save list of results
EACvsWAC_WX_res <- list(
  statistic = wx.eac.vs.wac$statistic, 
  p_value = wx.eac.vs.wac$p.value)
                      
  ##Convert the list to a data frame
EACvsWAC_WX_res <- as.data.frame(EACvsWAC_WX_res)

  ##Save the data frame to a CSV file
write.table(EACvsWAC_WX_res, file = "EACvsWAC_WX_res.csv", row.names = TRUE, col.names = TRUE, sep = ",", dec = ".")

median(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "WAC"])
median(ratios.isoc.counts$On_isochronous_norm[ratios.isoc.counts$Subspecies == "EAC"])
```
Yes
