---
title: "Drumming regional variation: pdfas, repdfas, glmms statistics"
author: Vesta Eleuteri
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    cards: true
    code_folding: hide
bibliography: references.bib
csl: apa.csl
---

# Introduction

This document accompanies the manuscript "Chimpanzee drumming shows rhythmicity and subspecies variation" @vestaeleuteriChimpanzeeDrummingShows2024.

Here we tested for regional variation in drumming by comparing drumming structure across communities, across populations, and between subspecies.
We first used permuted discriminant function analyses (pDFA) to test for the group differences and then used repeated discriminant function analyses (repDFA)
and generalized linear mixed models (glmms) to identify the key variables discriminating the groups in significant pDFAs.

Here we also tested for regional variation in drumming behaviour by comparing the timing of drumming onset during pant-hoot vocalizations across communities,
across populations, and between subspecies. We used glmms to test for the group differences.

# Preliminaries

## Load packages, functions, data

```{r load}
# for easy loading of packages
if (!require("pacman")) install.packages("pacman")

# load packages
p_load("here", "devtools", "gobbios/cfp", "gobbios/cfp@v.0.1.0", "cfp", "MASS", "lmerTest", "lme4", "MuMIn", "dplyr", "knitr", "kableExtra")

here <- here::here

# for relative file paths (e.g. here('Input', 'file.csv'))
here::i_am(file.path("pdfa_repdfa_glmm.Rmd"))

# get custom R functions
source(here("functions", "pdfa_functions.r"))
source(here("functions", "repDFA_nested.r"))
source(here("functions", "diagnostic_fcns.r"))
source(here("functions", "glmm_stability.r"))
source(here("functions", "boot_glmm.r"))

# Knitr output settings
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  fig.width = 10,
  fig.height = 10,
  cache = TRUE
)

# load data
xdata <- read.csv(here("dataframes", "drumming_wide.csv"))
```

# Summary data

## Summary data community

```{r summary-com}
summary_stats_com <- xdata %>%
  group_by(Subspecies, Population, Community) %>%
  dplyr::summarise(
    Number_Drums = n(),
    Mean_N_hits = round(mean(N_hits, na.rm = TRUE), 3),
    SD_N_hits = round(sd(N_hits, na.rm = TRUE), 3),
    Mean_Bout_Duration = round(mean(Total_bout_duration, na.rm = TRUE), 3),
    SD_Bout_Duration = round(sd(Total_bout_duration, na.rm = TRUE), 3),
    Mean_Bout_cv = round(mean(Bout_cv, na.rm = TRUE), 3),
    SD_Bout_cv = round(sd(Bout_cv, na.rm = TRUE), 3),
    Mean_Bout_npvi = round(mean(Bout_npvi, na.rm = TRUE), 3),
    SD_Bout_npvi = round(sd(Bout_npvi, na.rm = TRUE), 3),
    Mean_Entropy = round(mean(Bout_entropy, na.rm = TRUE), 3),
    SD_Entropy = round(sd(Bout_entropy, na.rm = TRUE), 3),
    .groups = "drop"
  )
summary_stats_com
```

## Summary data population

```{r summary-pop}
summary_stats_pop <- xdata %>%
  group_by(Subspecies, Population) %>%
  dplyr::summarise(
    Number_Drums = n(),
    Mean_N_hits = round(mean(N_hits, na.rm = TRUE), 3),
    SD_N_hits = round(sd(N_hits, na.rm = TRUE), 3),
    Mean_Bout_Duration = round(mean(Total_bout_duration, na.rm = TRUE), 3),
    SD_Bout_Duration = round(sd(Total_bout_duration, na.rm = TRUE), 3),
    Mean_Bout_cv = round(mean(Bout_cv, na.rm = TRUE), 3),
    SD_Bout_cv = round(sd(Bout_cv, na.rm = TRUE), 3),
    Mean_Bout_npvi = round(mean(Bout_npvi, na.rm = TRUE), 3),
    SD_Bout_npvi = round(sd(Bout_npvi, na.rm = TRUE), 3),
    Mean_Entropy = round(mean(Bout_entropy, na.rm = TRUE), 3),
    SD_Entropy = round(sd(Bout_entropy, na.rm = TRUE), 3),
    .groups = "drop" # Drop the grouping after summarising
  )
summary_stats_pop
```

## Summary data subspecies

```{r summary-sub}
summary_stats_sub <- xdata %>%
  group_by(Subspecies) %>%
  dplyr::summarise(
    Number_Drums = n(),
    Mean_N_hits = round(mean(N_hits, na.rm = TRUE), 3),
    SD_N_hits = round(sd(N_hits, na.rm = TRUE), 3),
    Mean_Bout_Duration = round(mean(Total_bout_duration, na.rm = TRUE), 3),
    SD_Bout_Duration = round(sd(Total_bout_duration, na.rm = TRUE), 3),
    Mean_Bout_cv = round(mean(Bout_cv, na.rm = TRUE), 3),
    SD_Bout_cv = round(sd(Bout_cv, na.rm = TRUE), 3),
    Mean_Bout_npvi = round(mean(Bout_npvi, na.rm = TRUE), 3),
    SD_Bout_npvi = round(sd(Bout_npvi, na.rm = TRUE), 3),
    Mean_Entropy = round(mean(Bout_entropy, na.rm = TRUE), 3),
    SD_Entropy = round(sd(Bout_entropy, na.rm = TRUE), 3),
    .groups = "drop" # Drop the grouping after summarising
  )
summary_stats_sub
```

## Summary data all

```{r summary-all}
summary_stats_all <- xdata %>%
  dplyr::summarise(
    Number_Drums = n(),
    Mean_N_hits = round(mean(N_hits, na.rm = TRUE), 3),
    SD_N_hits = round(sd(N_hits, na.rm = TRUE), 3),
    Mean_Bout_Duration = round(mean(Total_bout_duration, na.rm = TRUE), 3),
    SD_Bout_Duration = round(sd(Total_bout_duration, na.rm = TRUE), 3),
    Mean_Bout_cv = round(mean(Bout_cv, na.rm = TRUE), 3),
    SD_Bout_cv = round(sd(Bout_cv, na.rm = TRUE), 3),
    Mean_Bout_npvi = round(mean(Bout_npvi, na.rm = TRUE), 3),
    SD_Bout_npvi = round(sd(Bout_npvi, na.rm = TRUE), 3),
    Mean_Entropy = round(mean(Bout_entropy, na.rm = TRUE), 3),
    SD_Entropy = round(sd(Bout_entropy, na.rm = TRUE), 3),
    .groups = "drop" # Drop the grouping after summarising
  )
summary_stats_all
```

# Analyses for regional variation in drumming structure

## Sort data

```{r drum-sort-data}
## Sort data
range(xdata$Total_bout_duration)
range(xdata$N_hits)
range(xdata$Bout_cv)
range(xdata$Bout_npvi)
range(xdata$Bout_entropy)

# Add constants to variables to logtransform all non-normal variables with 0 bounds
c <- 0.0001
xdata$Total_bout_duration_c <- xdata$Total_bout_duration + c
xdata$N_hits_c <- xdata$N_hits + c
xdata$Bout_cv_c <- xdata$Bout_cv + c
xdata$Bout_npvi_c <- xdata$Bout_npvi + c
xdata$Bout_entropy_c <- xdata$Bout_entropy + c

# Logtransform skewed distributions
hist(xdata$N_hits_c) # skewed
hist(log(xdata$N_hits_c)) # better
xdata$log.N_hits_c <- log(xdata$N_hits_c) # logtransform

hist(xdata$Total_bout_duration_c) # skewed
hist(log(xdata$Total_bout_duration_c)) # better
xdata$log.Total_bout_duration_c <- log(xdata$Total_bout_duration_c) # logtransform

hist(xdata$Bout_cv_c) # slightly skewed
hist(log(xdata$Bout_cv_c)) # skewed on other side so do not logtransform

hist(xdata$Bout_npvi_c) # slightly skewed
hist(log(xdata$Bout_npvi_c)) # skewed on other side so do not logtransform

hist(xdata$Bout_entropy_c) # slightly skewed
hist(log(xdata$Bout_entropy_c)) # skewed on other side so do not logtransform
```

## Community pDFA

We first tested for regional variation using a pDFA to compare drumming structure across communities.

```{r pdfa-community}
set.seed(131)
pdfa.res.communities <- pDFA.nested(
  test.fac = "Pop_com_code", contr.fac = "Com_indiv_code",
  variables = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ),
  restrict.by = NULL,
  n.contr.fac.levels.to.sel = NULL,
  n.to.sel.per.contr.fac.level = NULL, n.sel = 100, n.perm = 1000,
  pdfa.data = xdata
)
pdfa.res.communities$result
```
The pDFA is significant showing that drumming structure varies across communities. We thus used a repDFA to identify the key variables significantly discriminating the communities.

### Communtiy repDFA

```{r repdfa-community}
set.seed(5)
ImpVars_all_comms_allvabs <- repDFA_nested(xdata,
  testfactor = "Pop_com_code",
  balancefactor = c("Pop_com_code", "Com_indiv_code"),
  varnames = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ),
  npercomb = 3, nrand = 1000
)

All_comms_allvabs_repDFA_res_df1 <- table(ImpVars_all_comms_allvabs$df1_best) # Bout_npvi_c=524 (52% of 100%) and N hits 460
All_comms_allvabs_repDFA_res_df1
All_comms_allvabs_repDFA_res_df2 <- table(ImpVars_all_comms_allvabs$df2_best) # log.Total_bout_duration_c=543, Bout_entropy_c=226
All_comms_allvabs_repDFA_res_df2
```
The repDFA identified as key variables on discriminant 1: nPVI and Number of hits, on discriminant 2: total bout duration and entropy. We thus used glmms to test if these variables significantly vary across the communities.

### Community glmm nPVI

```{r glmm-community-nPVI}
## Run model
Model_npvi <- lmer(Bout_npvi_c ~ Community +
  (1 | Individual), data = xdata, REML = F)

## Full-null model comparison
null_Model_npvi <- lmer(Bout_npvi_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_npvi <- as.data.frame(anova(null_Model_npvi, Model_npvi, test = "Chisq")) # sig
Chisq_Model_npvi <- round(Chisq_Model_npvi, 3)
Chisq_Model_npvi
```
nPVI significantly varies across communities.

### Community glmm N hits

```{r glmm-community-N hits}
## Run model
Model_numhits <- lmer(log.N_hits_c ~ Community +
  (1 | Individual), data = xdata, REML = F)

## Full-null model comparison
null_Model_numhits <- lmer(log.N_hits_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_numhits <- as.data.frame(anova(null_Model_numhits, Model_numhits, test = "Chisq")) # sig
Chisq_Model_numhits <- round(Chisq_Model_numhits, 3)
Chisq_Model_numhits
```
N hits significantly varies across communities.

### Community glmm bout duration

```{r glmm-community-bout-duration}
## Run model
Model_drumduration <- lmer(log.Total_bout_duration_c ~ Community +
  (1 | Individual), data = xdata, REML = F)

## Full-null model comparison
null_Model_drumduration <- lmer(log.Total_bout_duration_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_drumduration <- as.data.frame(anova(null_Model_drumduration, Model_drumduration, test = "Chisq"))
Chisq_Model_drumduration
```
Bout duration does not significantly vary across communities.

### Community glmm entropy

```{r glmm-community-entropy}
## Run model
Model_entropy <- lmer(Bout_entropy_c ~ Community +
  (1 | Individual), data = xdata, REML = F)

## Full-null model comparison
null_Model_entropy <- lmer(Bout_entropy_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_entropy <- as.data.frame(anova(null_Model_entropy, Model_entropy, test = "Chisq"))
Chisq_Model_entropy <- round(Chisq_Model_entropy, 3)
Chisq_Model_entropy
```
Entropy does not significantly vary across communities.

## Confirmatory Subspecies pDFA

Because plots (see main_plots.Rmd) suggest that the variation lies across communities of different subspecies, we followed up the community comparison with a pDFA comparing the two subspecies (eastern and western).

```{r pdfa-subspecies}
set.seed(126)
pdfa.res.subspecies <- pDFA.nested(
  test.fac = "Subspecies", contr.fac = "Com_indiv_code",
  variables = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ), restrict.by = NULL,
  n.contr.fac.levels.to.sel = NULL,
  n.to.sel.per.contr.fac.level = NULL, n.sel = 100, n.perm = 1000,
  pdfa.data = xdata
)


pdfa.res.subspecies$result
```
The pDFA is significant showing that drumming structure varies between subspecies. We thus used a repDFA to identify the key variables significantly discriminating the subspecies.

### Subspecies repDFA

```{r repdfa-subspecies}
set.seed(5)
ImpVars_sub_allvabs <- repDFA_nested(xdata,
  testfactor = "Subspecies",
  balancefactor = c("Subspecies", "Com_indiv_code"),
  varnames = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ),
  npercomb = 3, nrand = 1000
)

Sub_allvabs_repDFA_res_df1 <- table(ImpVars_sub_allvabs$df1_best)
Sub_allvabs_repDFA_res_df1
```
The repDFA idenfitied as most contributing variables on discriminant 1: nPVI and Number of hits. We thus used glmms to test if these variables significantly discriminate the subspecies.

### Subspecies glmm nPVI

```{r glmm-subspecies-nPVI}
## Run model
Model_npvi_sub <- lmer(Bout_npvi_c ~ Subspecies +
  (1 | Individual), data = xdata, REML = F)

## Full-null model comparison
null_Model_npvi_sub <- lmer(Bout_npvi_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_npvi_sub <- as.data.frame(anova(null_Model_npvi_sub, Model_npvi_sub, test = "Chisq")) # sig
Chisq_Model_npvi_sub <- round(Chisq_Model_npvi_sub, 3)
Chisq_Model_npvi_sub
```
nPVI significantly varies between subspecies.

### Subspecies glmm N hits

```{r glmm-subspecies-N hits}
## Run model
Model_nhits_sub <- lmer(log.N_hits_c ~ Subspecies +
  (1 | Individual), data = xdata, REML = F)


## Full-null model comparison
null_Model_nhits_sub <- lmer(log.N_hits_c ~ (1 | Individual), data = xdata, REML = F)
Chisq_Model_nhits_sub <- as.data.frame(anova(null_Model_nhits_sub, Model_nhits_sub, test = "Chisq")) # sig
Chisq_Model_nhits_sub <- round(Chisq_Model_nhits_sub, 3)
Chisq_Model_nhits_sub
```
N hits also significantly varies between subspecies.

Taken together, our results above show that -across both communities and subspecies- western chimpanzees drum with lower nPVI (i.e., variability between adjacent pairs of inter-hit intervals) and more hits in drumming bouts that eastern chimpanzees.

## Across subspecies Populations pDFA

Because the sampled populations live in different habitat types, we used another pDFA to test whether ecology might affect drumming by comparing drumming structure across populations.

```{r pdfa-populations-all}
set.seed(128)
pdfa.res.populations <- pDFA.nested(
  test.fac = "Population", contr.fac = "Com_indiv_code",
  variables = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ), restrict.by = NULL,
  n.contr.fac.levels.to.sel = NULL,
  n.to.sel.per.contr.fac.level = NULL, n.sel = 100, n.perm = 1000,
  pdfa.data = xdata
)


pdfa.res.populations$result
```
The pDFA is non-significant showing that drumming structure does not vary across all populations and thus suggesting that ecology may not impact drumming structure.
But because the populations belong to different subspecies, we further assessed whether ecology may affect drumming structure while controlling for subspecies variation. We used two additional pDFAs comparing populations within each subspecies.

## Western populations pDFA

```{r pdfa-populations-west}
xdata_TNF <- subset(xdata, Population %in% c("Tai", "Nimba", "Fongoli"))
xdata_TNF <- droplevels(xdata_TNF)

set.seed(136)
pdfa.res.TNFpopulations <- pDFA.nested(
  test.fac = "Population", contr.fac = "Com_indiv_code",
  variables = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ), restrict.by = NULL,
  n.contr.fac.levels.to.sel = NULL,
  n.to.sel.per.contr.fac.level = NULL, n.sel = 100, n.perm = 1000,
  pdfa.data = xdata_TNF
)


pdfa.res.TNFpopulations$result
```
The pDFA is non-significant showing that drumming structure does not vary across western populations.

### Eastern populations pDFA

```{r pdfa-populations-east}
xdata_BKG <- subset(xdata, Population %in% c("Budongo", "Kibale", "Gombe"))
xdata_BKG <- droplevels(xdata_BKG)

set.seed(137)
pdfa.res.BKGpopulations <- pDFA.nested(
  test.fac = "Population", contr.fac = "Com_indiv_code",
  variables = c(
    "log.Total_bout_duration_c", "log.N_hits_c",
    "Bout_cv_c", "Bout_npvi_c", "Bout_entropy_c"
  ), restrict.by = NULL,
  n.contr.fac.levels.to.sel = NULL,
  n.to.sel.per.contr.fac.level = NULL, n.sel = 100, n.perm = 1000,
  pdfa.data = xdata_BKG
)


pdfa.res.BKGpopulations$result
```
The pDFA is non-significant showing that drumming structure does not vary across eastern populations.

# Analyses for regional variation in drumming onset during pant-hoot

We last tested for regional variation in drumming behaviour using 3 glmms to compare the timing of drumming onset during pant-hoot vocalizations across communities, across populations, and between subspecies.

## Sort data

```{r drum-ph-sort-data}
levels(xdata$With_Ph)[levels(xdata$With_Ph) %in% c("with")] <- "With"
xdata_ph <- subset(xdata, Start_Ph %in% c("Build-Up", "Climax", "Let-Down", "roar", "Roar")) # keeping only drums starting in clear pant-hoot phases
xdata_ph <- droplevels(xdata_ph)
xdata_ph$Start_Ph <- factor(xdata_ph$Start_Ph) # conver to factor
levels(xdata_ph$Start_Ph)[levels(xdata_ph$Start_Ph) %in% c("Build-Up", "roar", "Roar")] <- "Before Climax Start" # rename level
levels(xdata_ph$Start_Ph)[levels(xdata_ph$Start_Ph) %in% c("Climax", "Let-Down")] <- "After Climax Start" # rename level

## Check frequencies of response variable
table(xdata_ph$Start_Ph, xdata_ph$Community) # South has 0 drums After Climax Start so remove level
xdata_ph <- subset(xdata_ph, Community != c("South")) # removed, n=302
xdata_ph <- droplevels(xdata_ph) # drop levels no observations
table(xdata_ph$Start_Ph, xdata_ph$Community)

## Dummy code response variable
xdata_ph$Start_Ph <- as.numeric(xdata_ph$Start_Ph == "Before Climax Start") # Dummy coded Before Climax Start as 1
```

## Community glmm

```{r drum-ph-com-glmm}
## Run model
Model_start_ph_com <- glmer(Start_Ph ~ Community +
  (1 | Individual), data = xdata_ph, family = binomial)

## Full-null model comparison
null_Model_start_ph_com <- glmer(Start_Ph ~
  (1 | Individual), data = xdata_ph, family = binomial)
Chisq_Model_start_ph_com <- as.data.frame(anova(null_Model_start_ph_com, Model_start_ph_com, test = "Chisq")) # sig
Chisq_Model_start_ph_com <- round(Chisq_Model_start_ph_com, 3)
Chisq_Model_start_ph_com
```
The timing of drumming onset significantly varies across communities.

## Population glmm
```{r drum-ph-pop-glmm}
## Run model
Model_start_ph_pop <- glmer(Start_Ph ~ Population +
  (1 | Individual), data = xdata_ph, family = binomial)

## Full-null model comparison
null_Model_start_ph_pop <- glmer(Start_Ph ~
  (1 | Individual), data = xdata_ph, family = binomial)
Chisq_Model_start_ph_pop <- as.data.frame(anova(null_Model_start_ph_pop, Model_start_ph_pop, test = "Chisq"))
Chisq_Model_start_ph_pop <- round(Chisq_Model_start_ph_pop, 3)
Chisq_Model_start_ph_pop
```
The timing of drumming onset significantly varies across all populations.

## Confirmatory subspecies glmm

```{r drum-ph-sub-glmm}
## Run model
Model_start_ph_sub <- glmer(Start_Ph ~ Subspecies +
  (1 | Individual), data = xdata_ph, family = binomial)

## Full-null model comparison
null_Model_start_ph_sub <- glmer(Start_Ph ~
  (1 | Individual), data = xdata_ph, family = binomial)
Chisq_Model_start_ph_sub <- as.data.frame(anova(null_Model_start_ph_sub, Model_start_ph_sub, test = "Chisq"))
Chisq_Model_start_ph_sub <- round(Chisq_Model_start_ph_sub, 3)
Chisq_Model_start_ph_sub
```
The timing of drumming onset significantly varies between subspecies.

We find that western chimpanzees start to drum earlier in the pant-hoot than eastern chimpanzees typically by drumming in the build-up phase.

# References
<div id="refs"></div>