---
title: "Drumming regional variation: plots"
author: Vesta Eleuteri & Jelle van der Werff
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    cards: true
bibliography: references.bib
csl: apa.csl
---

# Introduction

This document accompanies the manuscript "Chimpanzee drumming shows rhythmicity and subspecies variation" @vestaeleuteriChimpanzeeDrummingShows2024.

Here we make plots to accompany our regional variation analyses.

# Preliminaries

```{r load, warning=FALSE, message=FALSE}
# for easy loading of packages
if (!require("pacman")) install.packages("pacman")

p_load(
  "here", "ggplot2", "ggdist"
)

# for relative file paths (e.g. here('Input', 'file.csv'))
here <- here::here
here::i_am(file.path("plots.Rmd"))

# Knitr output settings
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  fig.width = 10,
  fig.height = 10,
  cache = TRUE
)

source(here("theme_drumming.R"))

# load data
xdata <- read.csv(here("dataframes", "drumming_wide.csv"))
xdata.ratios <- read.csv(here("dataframes", "drumming_ratios.csv"))
```

# Plot for nPVI by community

```{r sort-data, warning=FALSE}
sub_order <- c("Western", "Eastern")
pop_order <- c("Fongoli", "Nimba", "Taï", "Gombe", "Kibale", "Budongo")
com_order <- c(
  "Fongoli", "Tongbongbon", "Gahtoy", "East", "Middle", "North", "South",
  "Kasekela", "Kanyawara", "Sonso", "Waibira"
)

xdata$Subspecies <- factor(xdata$Subspecies, levels = sub_order)
xdata$Population <- factor(xdata$Population, levels = pop_order)
xdata$Community <- factor(xdata$Community, levels = com_order)

# sort ratios
xdata.ratios$Subspecies <- factor(xdata.ratios$Subspecies, levels = sub_order)
xdata.ratios$Population <- factor(xdata.ratios$Population, levels = pop_order)
xdata.ratios$Community <- factor(xdata.ratios$Community, levels = com_order)

# Create a new label combining Community and Population for plots
xdata$CombinedLabel <- paste(xdata$Population, "\n", xdata$Community)


desired_order <- c(
  "Fongoli \n Fongoli", "Nimba \n Tongbongbon", "Nimba \n Gahtoy",
  "Taï \n East", "Taï \n Middle", "Taï \n North", "Taï \n South",
  "Gombe \n Kasekela", "Kibale \n Kanyawara", "Budongo \n Sonso", "Budongo \n Waibira"
)


# Convert CombinedLabel to a factor with the specified order
xdata$CombinedLabel <- factor(xdata$CombinedLabel, levels = desired_order)
```

```{r nPVI-plot-com, fig.width=26}
Com_npvi_plot <- ggplot(xdata, aes(x = factor(CombinedLabel), y = Bout_npvi, fill = factor(CombinedLabel))) +
  # add half-violin from {ggdist} package
  stat_halfeye(
    # adjust bandwidth
    adjust = 0.5,
    # move to the right
    justification = -0.2,
    # remove the slub interval
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  geom_point(
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  # Themes and Labels
  theme_drumming() + # use drumming plot theme
  labs(
    title = "D - Community variability between adjacent pairs of inter-hit intervals",
    x = "",
    y = "nPVI",
    fill = ""
  ) +
  guides(fill = "none") +
  theme(
    plot.title = element_text(size = 40),
    axis.title.x = element_text(size = 34, margin = margin(t = 20)),
    axis.title.y = element_text(size = 34),
    axis.text.x = element_text(size = 30, margin = margin(t = 20)), # Add margin to x-axis text
    axis.text.y = element_text(size = 30),
    legend.title = element_text(size = 34),
    legend.text = element_text(size = 34)
  ) +
  theme(plot.margin = margin(1, 0, 2, 1, "cm"))

Com_npvi_plot
ggsave(here("plots", "Com_nPVI.png"), width = 26, height = 8)
ggsave(here("plots", "Com_nPVI.svg"), width = 26, height = 8)
```

# Plot for drumming onset in pant-hoot

```{r sort-ph-data, warning=FALSE}
xdata_ph <- subset(xdata, Start_Ph %in% c("Build-Up", "Climax", "Let-Down", "roar", "Roar"))
xdata_ph <- droplevels(xdata_ph)
xdata_ph$Start_Ph <- factor(xdata_ph$Start_Ph) # conver to factor
levels(xdata_ph$Start_Ph)[levels(xdata_ph$Start_Ph) %in% c("Build-Up", "roar", "Roar")] <- "Before climax start"
levels(xdata_ph$Start_Ph)[levels(xdata_ph$Start_Ph) %in% c("Climax", "Let-Down")] <- "After climax start"
xdata_ph <- droplevels(xdata_ph)

community_phdrumstart_table <- table(xdata_ph$CombinedLabel, xdata_ph$Start_Ph)

community_phdrumstart_data <- as.data.frame(as.table(community_phdrumstart_table))
colnames(community_phdrumstart_data) <- c("Community", "Drumming_Start", "Frequency")

### Convert Frequency to Percentage
community_phdrumstart_data$Percentage <- with(
  community_phdrumstart_data,
  ave(Frequency, Community, FUN = function(x) 100 * x / sum(x))
)
```


```{r drum-ph-plot-com, fig.width=26}
colors <- c("white", "black")
Com_ph_start_plot <- ggplot(community_phdrumstart_data, aes(x = Community, y = Percentage, fill = Drumming_Start)) +
  geom_bar(stat = "identity", colour = "black") + # Add border around bars
  theme_drumming() +
  scale_fill_manual(values = colors) +
  labs(title = "E - Community drumming onset during pant-hoot", x = "", y = "Percentage", fill = "") +
  theme(
    plot.title = element_text(size = 40),
    axis.title.x = element_text(size = 34, margin = margin(t = 20)),
    axis.title.y = element_text(size = 34),
    axis.text.x = element_text(size = 30),
    axis.text.y = element_text(size = 30),
    legend.title = element_text(size = 34),
    legend.text = element_text(size = 34)
  ) +
  theme(plot.margin = margin(1, 0, 2, 1, "cm")) +
  guides(fill = "none")

Com_ph_start_plot
ggsave(here("plots", "Com_PhStart.png"), width = 26, height = 8)
ggsave(here("plots", "Com_PhStart.svg"), width = 26, height = 8)
```

# Plot for number of hits in drumming bout

```{r N hits-plot-sub, fig.width=26}
Sub_nhits_plot <- ggplot(xdata, aes(x = N_hits, fill = factor(Subspecies))) +
  geom_histogram(
    position = "identity",
    width = 0.5,
    alpha = 0.7,
    bins = 28 # Number of bins
  ) +
  # Themes and Labels
  theme_drumming() + # Use drumming plot theme
  labs(
    title = "Subspecies number of hits in drumming bout",
    x = "Number of hits",
    y = "Frequency",
    fill = "Subspecies"
  ) +
  scale_x_continuous(
    breaks = c(1, 2, seq(3, 28, by = 1)), # Add breaks starting from 1 to 28
    limits = c(1, 28) # adjust limits on x
  ) +
  theme(
    plot.title = element_text(size = 40),
    axis.title.x = element_text(size = 34),
    axis.title.y = element_text(size = 34),
    axis.text.x = element_text(size = 30),
    axis.text.y = element_text(size = 30),
    legend.title = element_text(size = 34, hjust = 1),
    legend.text = element_text(size = 34, hjust = 1),
    legend.position = "right", # Position the legend
    legend.key.size = unit(1, "cm"), # Adjust size of legend keys
    legend.spacing.y = unit(0.5, "cm") # Adjust spacing between legend items
  ) +
  theme(plot.margin = margin(1, 0, 2, 1, "cm"))
Sub_nhits_plot

ggsave(here("plots", "Sub_Num_Hits.pdf"), width = 25, height = 10)
ggsave(here("plots", "Sub_Num_Hits.png"), width = 25, height = 10)
```
# References

<div id="refs"></div>