---
title: "Drumming chimps: Ratios statistics"
author: Vesta Eleuteri & Jelle van der Werff
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    cards: true
    code_folding: hide
bibliography: references.bib
csl: apa.csl
---

# Introduction

This document accompanies the manuscript "Chimpanzee drumming shows rhythmicity and subspecies variation" @vestaeleuteriChimpanzeeDrummingShows2024.



# Preliminaries

## Load packages and data

```{r load-packages, echo = FALSE, warning = FALSE}
# for easy loading of packages
if (!require("pacman")) install.packages("pacman")

# load packages
p_load("here", "ggplot2", "dplyr", "nlme", "dplyr", "plyr", "knitr", "rmdformats", "kableExtra", "Matching", "kldtools")

here <- here::here

# for relative file paths (e.g. here('Input', 'file.csv'))
here::i_am(file.path("ratios.Rmd"))

# load data
ratios <- read.csv(here("dataframes", "drumming_ratios.csv"))
ibis <- read.csv(here("dataframes", "drumming_long.csv"))

# load custom ggplot theme
source(here("theme_drumming.R"))

# create simulated data based on uniform distribution
ratios.simulated <- runif(n = 1000001, min = min(ibis$IBI_ms), max = max(ibis$IBI_ms)) # IOIs

# calculate ratios
ratios.simulated <- ratios.simulated / (ratios.simulated + lead(ratios.simulated))
ratios.simulated <- ratios.simulated[!is.na(ratios.simulated)]
ratios.simulated <- as.data.frame(list(Ratio = ratios.simulated))

# set knitr options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, out.width = "100%", fig.height = 10, fig.width = 10, comment = NA)
```


## Defining bins and coding ratios categories

```{r bins}
bins <- list(
  "1" = c(0, 1 / 4.25), # 1 / 4.5
  "2" = c(1 / 4.25, 1 / 3.75), # 1 / 4 > 1:3
  "3" = c(1 / 3.75, 1 / 3.25), # 1 / 3.5
  "4" = c(1 / 3.25, 1 / 2.75), # 1 / 3 > 1:2
  "5" = c(1 / 2.75, 1 / 2.25), # 1 / 2.5
  "6" = c(1 / 2.25, 1 - 1 / 2.25), # 1 / 2 > 1:1
  "7" = c(1 - 1 / 2.25, 1 - 1 / 2.75), # 1 - 1 / 2.5
  "8" = c(1 - 1 / 2.75, 1 - 1 / 3.25), # 1 - 1 / 3 > 2:1
  "9" = c(1 - 1 / 3.25, 1 - 1 / 3.75), # 1 - 1 / 3.5
  "10" = c(1 - 1 / 3.75, 1 - 1 / 4.25), # 1 - 1 / 4 > 3:1
  "11" = c(1 - 1 / 4.25, 1) # 1 - 1 / 4.5
)

breaks <- c(0, 1 / 4.25, 1 / 3.75, 1 / 3.25, 1 / 2.75, 1 / 2.25, 1 - 1 / 2.25, 1 - 1 / 2.75, 1 - 1 / 3.25, 1 - 1 / 3.75, 1 - 1 / 4.25, 1)

labels <- names(bins)

ratios$Bin <- cut(ratios$Ratio, breaks = breaks, labels = labels)
ratios.simulated$Bin <- cut(ratios.simulated$Ratio, breaks = breaks, labels = labels)
```

## Reordering/renaming variables etc.

```{r reorder}
# Change levels order
ibis$Subspecies <- factor(ibis$Subspecies, levels = c("Western", "Eastern"))
ibis$Population <- factor(ibis$Population, levels = c("Fongoli", "Nimba", "Taï", "Gombe", "Kibale", "Budongo"))
ibis$Community <- factor(ibis$Community, levels = c(
  "Fongoli", "Tongbongbon", "Gahtoy", "East", "Middle", "North", "South",
  "Kasekela", "Kanyawara", "Sonso", "Waibira"
))

ratios$Subspecies <- factor(ratios$Subspecies, levels = c("Western", "Eastern"))
ratios$Population <- factor(ratios$Population, levels = c("Fongoli", "Nimba", "Taï", "Gombe", "Kibale", "Budongo"))
ratios$Community <- factor(ratios$Community, levels = c(
  "Fongoli", "Tongbongbon", "Gahtoy", "East", "Middle", "North", "South",
  "Kasekela", "Kanyawara", "Sonso", "Waibira"
))
```

# Descriptives (inter-hit interval and ratios)

```{r descriptives}
subspecies.table <- ddply(ibis, .(Subspecies), summarize,
  mean = mean(IBI_ms),
  median = median(IBI_ms),
  sd = sd(IBI_ms),
  min = min(IBI_ms),
  max = max(IBI_ms)
)
kable(subspecies.table, format = "html", caption = "Descriptives of inter-hit intervals (ms) by subspecies") %>%
  kable_styling("striped", full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:6, width = "100px")

overall.table <- as.data.frame(ddply(ibis, .(), summarize,
  mean = mean(IBI_ms),
  median = median(IBI_ms),
  sd = sd(IBI_ms),
  min = min(IBI_ms),
  max = max(IBI_ms)
))
overall.table <- overall.table[, 2:6]

kable(overall.table, format = "html", caption = "Descriptives of inter-hit intervals (ms) overall") %>%
  kable_styling("striped", full_width = FALSE)
```

# Main text: Statistics

## Introduction

Our main analysis relies on the method devised by @roeskeCategoricalRhythmsAre2020, in which the relationships between adjacent inter-onset/inter-hit intervals
(IOIs/IHIs) are described by the formula: $r_k = \frac{IOI_k}{IOI_k + IOI_{k+1}}$. This formula results in ratios $r_k$ that range from 0 to 1, where e.g. 0.5
indicates isochrony.

Recently, @jadoulHiddenAssumptionsInteger2024 have highlighted a number of issues with the usual method of doing statistics with these ratios, and have improved the
method as hitherto used. Our analysis takes into account these new insights.

Classically, the null hypothesis is that observed IOIs are distributed according to a random uniform distribution. However, the statistical tests that are
generally used do not conform to this null hypothesis [@jadoulHiddenAssumptionsInteger2024]. According to @jadoulHiddenAssumptionsInteger2024, the assumption of a uniformly random
distribution of IOIs as the null-hypothesis means that the ratios are distributed according to the distribution plotted below. The exact shape of this
distribution is dependent on the left and right bounds of the random uniform distribution used for sampling the IOIs. However, the shape will remain similar
to the one plotted below.

As left (a) and right (b) bounds of the uniform distribution of IOIs, in our analysis we use the minimum observed IOI duration and the maximum observed IOI
duration, respectively.

The dotted lines in the plot are the histogram bins, following @roeskeCategoricalRhythmsAre2020, which represent the integer ratios 1:3 (bin centered around 0.25),
1:2 (bin centered around 0.33), 1:1 (isochrony; bin centered around 0.5), 2:1 (bin centered around 0.66), and 3:1 (bin centered around 0.75).
The bins in between these integer ratios are the off-integer bins.

```{r uniform-random-distribution}
ggplot(ratios.simulated, aes(x = Ratio)) +
  geom_density(linewidth = 2) +
  theme_drumming() +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.5, colour = "grey") +
  labs(x = "Ratio", y = "Probability density", title = "Random ratios distribution", subtitle = "Assumes uniformly random IOIs") +
  scale_x_continuous(
    breaks = c(0, 0.25, 0.3333, 0.5, 0.6666, 0.75, 1), limits = c(0, 1),
    labels = c("0.0", "0.25\n1:3", "0.33\n1:2", "0.5\n1:1", "0.66\n3:2", "0.75\n3:1", "1.0")
  )
```

We define the bins following Roeske et al. (2020):

```{r surface-areas}
ratios.bins <- data.frame(bin = as.character(1:11))

ratios.bins$bin_left <- sapply(bins, function(x) x[1])
ratios.bins$bin_right <- sapply(bins, function(x) x[2])
ratios.bins$bin_name <- c("1/4.5", "1/4", "1/3.5", "1/3", "1/2.5", "1/2", "1 - 1/2", "1 - 1/3", "1 - 1/3.5", "1 - 1/4", "1 - 1/4.5")
ratios.bins$bin_alternative_name <- c("", "1:3", "", "1:2", "", "1:1", "", "2:1", "", "3:1", "")

kable(ratios.bins,
  format = "html", caption = "Bins definition", col.names =
    c("Bin", "Bin (left bound)", "Bin (right bound)", "Bin center", "Integer ratio")
) %>%
  kable_styling("striped", full_width = FALSE)
```

Then, we count the number of observations in each bin.

```{r count-and-normalize-drumming}
# Code off- and on-isochronous
ratios$On_off_isoc <- NA
ratios$On_off_isoc <- factor(ratios$On_off_isoc, levels = c("Off", "On", "Other"))
ratios$On_off_isoc <- "Other"
ratios$On_off_isoc[ratios$Bin %in% c(5, 7)] <- "Off"
ratios$On_off_isoc[ratios$Bin == 6] <- "On"

# Count
ratios.isoc.counts <- dplyr::summarize(dplyr::group_by(ratios, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)

kable(head(ratios.isoc.counts, 10),
  format = "html", align = "l",
  caption = "Counts of on- and off-isochronous drumming ratios (first 10 rows)",
  col.names = c(
    "Subspecies", "Population", "Community", "Community (Individual)", "On-isochronous count", "Off-isochronous count", "Other ratios count",
    "Total n ratios observed"
  )
) %>%
  kable_styling("striped", full_width = FALSE)
```


## Plot of ratios distributions by subspecies

```{r plot-ratios-distribution}
labels <- as.data.frame(list(
  x = c(0.25, 0.335, 0.5, 0.665, 0.75),
  y = c(0.1, 0.1, 0.1, 0.1, 0.1),
  label = c("1:3", "1:2", "1:1", "2:1", "3:1"),
  size = c(4, 6, 10, 6, 4)
))

ggplot(ratios, aes(x = Ratio, fill = Subspecies)) +
  geom_density(linewidth = 0, alpha = 0.7) +
  geom_density(data = ratios.simulated, aes(x = Ratio), linewidth = 2, colour = "yellow", inherit.aes = FALSE, show.legend = FALSE) +
  theme_drumming() +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.5, colour = "grey") +
  labs(y = "Density", x = "r", fill = "", title = "A - Subspecies rhythmic patterns") +
  geom_label(data = labels, aes(x = x, y = y, label = label), size = 10, family = "Helvetica", inherit.aes = FALSE) +
  theme(
    plot.title = element_text(size = 38),
    axis.title.x = element_text(size = 32),
    axis.title.y = element_text(size = 32),
    axis.text.x = element_text(size = 28),
    axis.text.y = element_text(size = 28),
    legend.title = element_text(size = 32),
    legend.text = element_text(size = 32)
  ) +
  theme(plot.margin = margin(1, 0, 1, 1, "cm")) +
  guides(fill = "none")

ggsave(here("plots", "integer_ratios_vs_random.png"), width = 26, height = 8)
ggsave(here("plots", "integer_ratios_vs_random.svg"), width = 26, height = 8)
```


## Statistics

We here test:

1. Whether the drumming ratios are randomly distributed (i.e. whether the observed ratios are different from a random distribution of ratios).
2. Whether the distributions of Eastern and Western are different from each other.
3. Whether there are more on-isochronous ratios than expected by chance.
4. Whether there is more isochrony in Eastern vs. Western.


### Is the drumming random? (drumming ratios vs. random distribution)

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when IOIs are uniformly distributed (which we simulated above). We use Kolmogorov-Smirnov tests for the comparisons.

#### All observations (both East and West subspecies) vs. uniformly random

```{r ks-test}
ks.all <- ks.test(ratios$Ratio, ratios.simulated$Ratio, alternative = "two.sided")
ks.all
```

**Across subspecies, the drumming is not random.**

#### Eastern chimps vs. uniformly random

```{r ks-test-eac}
ks.eac <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios.simulated$Ratio, alternative = "two.sided")
ks.eac
```

**For Eastern chimps, the drumming is not random.**

#### Western chimps vs. uniformly random

```{r ks-test-wac}
ks.wac <- ks.test(ratios$Ratio[ratios$Subspecies == "Western"], ratios.simulated$Ratio, alternative = "two.sided")
ks.wac
```

**For Western chimps, the drumming is not random.**

### Are the distributions of Eastern and Western different from each other?

```{r ks-test-eac-vs-wac}
ks.eac.vs.wac <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios$Ratio[ratios$Subspecies == "Western"], alternative = "two.sided")
ks.eac.vs.wac
```

**The distributions of Eastern and Western are different from each other.**

### Are there more on-isochronous ratios than expected by chance?

Here we compare the proportion of on-isochronous ratios to the proportion expected by chance (assuming uniformly random IOIs).

We do so by subtracting the surface area of the isochronous bin from each empirical probability
(i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) per individual.

Then we test (using a one-sample Wilcoxon signed-rank test) whether those values are different from 0 (where 0 indicates chance level).

The surface area of the isochronous bin is here calculated using the formula provided by @jadoulHiddenAssumptionsInteger2024:

$$\frac{1}{2} - \frac{(4b - 5a)^2}{40 (b - a)^2}$$

```{r isoc-vs-random-preparation}
# Calculate the surface area of the isochronous bin using the formula from Jadoul et al. (2024)
a <- min(ibis$IBI_ms)
b <- max(ibis$IBI_ms)
surface_area_isochronous <- ((1 / 2) - (((4 * b - 5 * a)^2) / (40 * (b - a)^2)))

# Multiply by two (there's one bin on each side of 0.5 when using this formula)
surface_area_isochronous <- surface_area_isochronous * 2 # 0.2042091

# Subtract surface area of isochronous bin from empirical probabilities
ratios.isoc.counts$Isoc_prob <- ratios.isoc.counts$On_isochronous / ratios.isoc.counts$Total_obs
ratios.isoc.counts$Isoc_prob_normalized <- ratios.isoc.counts$Isoc_prob - surface_area_isochronous
```

```{r isoc-vs-random-plot}
ggplot(ratios.isoc.counts, aes(y = Isoc_prob * 100, x = Subspecies, fill = Subspecies)) +
  geom_boxplot(colour = "black", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, show.legend = FALSE) +
  theme_drumming() +
  labs(
    y = "", x = "",
  ) +
  geom_hline(yintercept = surface_area_isochronous * 100, linetype = "dashed", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

ggsave(here("plots", "isochronous_ratios_vs_random.svg"), width = 4, height = 4)
ggsave(here("plots", "isochronous_ratios_vs_random.png"), width = 4, height = 4)
```


#### All observations

```{r isoc-vs-random-wilcoxon-all}
isoc.vs.random.wsrt.all <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized, alternative = "two.sided")
isoc.vs.random.wsrt.all
print(paste("Median for all chimps:", median(ratios.isoc.counts$Isoc_prob_normalized)))
```

**Across subspecies, there are not more on-isochronous ratios than expected by chance.**

#### Eastern chimps

```{r isoc-vs-random-wilcoxon-eac}
isoc.vs.random.wsrt.eac <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Eastern"],
  alternative = "two.sided"
) # Significant
isoc.vs.random.wsrt.eac
print(paste("Median for Eastern:", median(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Eastern"])))
```

**In Eastern chimps, there are significantly _fewer_ on-isochronous ratios than expected by chance.**

#### Western chimps

```{r isoc-vs-random-wilcoxon-wac}
isoc.vs.random.wsrt.wac <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Western"],
  alternative = "two.sided"
) # Significant
isoc.vs.random.wsrt.wac
print(paste("Median for Western:", median(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Western"])))
```

**In Western chimps, there are significantly _more_ on-isochronous ratios than expected by chance.**

### Is there more isochrony in Eastern vs. Western?

Here we compare the empirical probabilities (i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) of on-isochronous
ratios between Eastern and Western chimps using a Wilcoxon rank-sum test.

```{r isoc-counts-wilcoxon-eac-vs-wac}
wilcox.test(
  ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Western"],
  ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Eastern"]
)
print(paste("Median for East:", median(ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Eastern"])))
print(paste("Median for West:", median(ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Western"])))
```

**Yes, there are more on-isochronous ratios in Western than in Eastern.**

# Supplement: Statistics (>= 5 hits)

## Preparations

We subset the data to only include drumming bouts with 5 or more hits.

```{r copy-df-and-filter-5hits}
ratios.5plushits <- ratios
ratios.5plushits <- ratios.5plushits %>%
  group_by(Drumming_bout) %>%
  filter(n() >= 3) %>% # remove drums < 3 ratios.5plushits (4 IBIs/5 beats)
  group_by(Com_indiv_code) %>% # remove individuals with <3 drums
  filter(n_distinct(Drumming_bout) >= 3) %>%
  group_by(Community) %>% # remove communities with <1 individual
  filter(n_distinct(Individual) > 1) %>%
  ungroup()
```

### Counting the number of empirical observations in each bin

Below, we count the number of empirical observations in each bin for bouts with 5 hits or more.

```{r count-and-normalize-drumming-5hits}
# Code off- and on-isochronous
ratios.5plushits$On_off_isoc <- NA
ratios.5plushits$On_off_isoc <- factor(ratios.5plushits$On_off_isoc, levels = c("Off", "On", "Other"))
ratios.5plushits$On_off_isoc <- "Other"
ratios.5plushits$On_off_isoc[ratios.5plushits$Bin %in% c(5, 7)] <- "Off"
ratios.5plushits$On_off_isoc[ratios.5plushits$Bin == 6] <- "On"

# Count
ratios.5plushits.isoc.counts <- dplyr::summarize(dplyr::group_by(ratios.5plushits, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)
```

## Statistics

### Is the drumming random? (drumming ratios vs. random distribution)

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when IOIs are uniformly distributed. We use the Kolmogorov-Smirnov test for this purpose.

#### All observations (both East and West subspecies) vs. uniformly random

```{r ks-test-5hits}
ks.all <- ks.test(ratios.5plushits$Ratio, ratios.simulated$Ratio, alternative = "two.sided")
ks.all
```

**Across subspecies and looking only at drumming bouts with 5 or more hits, the drumming is not random.**

#### Eastern chimps vs. uniformly random

```{r ks-test-eac-5hits}
ks.eac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Eastern"], ratios.simulated$Ratio, alternative = "two.sided")
ks.eac
```

**For Eastern chimps, looking only at drumming bouts with 5 or more hits, the drumming is random.**

#### Western chimps vs. uniformly random

```{r ks-test-wac-5hits}
ks.wac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Western"], ratios.simulated$Ratio, alternative = "two.sided")
ks.wac
```

**For Western chimps, looking only at drumming bouts with 5 or more hits, the drumming is not random.**


### Are the distributions of Eastern and Western different from each other?

```{r ks-test-eac-vs-wac-5hits}
ks.eac.vs.wac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Eastern"],
  ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Western"],
  alternative = "two.sided"
)
ks.eac.vs.wac
```

**The distributions of Eastern and Western are different from each other, also when looking only at drumming bouts with 5 or more hits.**


### Are there more on-isochronous ratios than expected by chance?

Here we compare the proportion of on-isochronous ratios to the proportion expected by chance (assuming uniformly random IOIs).

We do so by subtracting the surface area of the isochronous bin from each empirical probability
(i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) per individual.

Then we test (using a one-sample Wilcoxon signed-rank test) whether those values are different from 0 (where 0 indicates chance level).

The surface area of the isochronous bin is here calculated using the formula provided by @jadoulHiddenAssumptionsInteger2024:

$$\frac{1}{2} - \frac{(4b - 5a)^2}{40 (b - a)^2}$$

```{r isoc-vs-random-preparation-5hits}
# Calculate the surface area of the isochronous bin using the formula from Jadoul et al. (2024)
a <- min(ibis$IBI_ms)
b <- max(ibis$IBI_ms)
surface_area_isochronous <- ((1 / 2) - (((4 * b - 5 * a)^2) / (40 * (b - a)^2)))

# Multiply by two (there's one bin on each side of 0.5 when using this formula)
surface_area_isochronous <- surface_area_isochronous * 2 # 0.2042091

# Subtract surface area of isochronous bin from empirical probabilities
ratios.5plushits.isoc.counts$Isoc_prob <- ratios.5plushits.isoc.counts$On_isochronous / ratios.5plushits.isoc.counts$Total_obs
ratios.5plushits.isoc.counts$Isoc_prob_normalized <- ratios.5plushits.isoc.counts$Isoc_prob - surface_area_isochronous
```

```{r isoc-vs-random-plot-5hits}
ggplot(ratios.5plushits.isoc.counts, aes(y = Isoc_prob * 100, x = Subspecies, fill = Subspecies)) +
  geom_boxplot(colour = "black", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, show.legend = FALSE) +
  theme_drumming() +
  labs(
    y = "Percentage of isochronous drumming ratios per individual", x = "Subspecies",
    title = "Percentage of isochronous ratios compared to chance level",
    subtitle = "Chance level indicated by dotted line is based on uniformly randomly distributed IOIs.\nOnly drumming bouts with 5 or more hits are included.",
  ) +
  geom_hline(yintercept = surface_area_isochronous * 100, linetype = "dashed", linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))
```

#### All observations

```{r isoc-vs-random-wilcoxon-all-5hits}
isoc.vs.random.wsrt.all <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized, alternative = "two.sided")
isoc.vs.random.wsrt.all
print(paste("Median for all chimps:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized)))
```

**Across subspecies, there are not more on-isochronous ratios than expected by chance, also when looking only at drumming bouts with 5 or more hits.**

#### Eastern chimps

```{r isoc-vs-random-wilcoxon-eac-5hits}
isoc.vs.random.wsrt.eac <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Eastern"], alternative = "two.sided") # Significant
isoc.vs.random.wsrt.eac
print(paste("Median for Eastern:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Eastern"])))
```

**In Eastern chimps, when looking only at drumming bouts with 5 or more events, there are not significantly more or fewer on-isochronous ratios than expected by chance.**

#### Western chimps

```{r isoc-vs-random-wilcoxon-wac-5hits}
isoc.vs.random.wsrt.wac <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Western"], alternative = "two.sided") # Significant
isoc.vs.random.wsrt.wac
print(paste("Median for Western:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Western"])))
```

**In Western chimps, there are significantly _more_ on-isochronous ratios than expected by chance, also when only looking at drumming bouts with 5 or more events.**


### Is there more isochrony in Eastern vs. Western?

Here we compare the empirical probabilities (i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) of on-isochronous
ratios between Eastern and Western chimps using a Wilcoxon rank-sum test.

```{r isoc-counts-wilcoxon-eac-vs-wac-5hits}
wilcox.test(
  ratios.5plushits.isoc.counts$Isoc_prob[ratios.5plushits.isoc.counts$Subspecies == "Western"],
  ratios.5plushits.isoc.counts$Isoc_prob[ratios.5plushits.isoc.counts$Subspecies == "Eastern"]
)
```

**Yes, there are more on-isochronous ratios in Western than in Eastern chimps, also when looking only at drumming bouts with 5 or more hits.**

# Supplement: Statistics (old method using bin width normalization)

## Introduction
We here use the method as has been used until now in the literature, which, due to the normalization procedure
(i.e. dividing by the bin _width_ rather than the bin _surface area_), assumes exponentially distributed IOIs resulting from a Poisson point-process
[@jadoulHiddenAssumptionsInteger2024].

## Define bins, bin widths, etc.

```{r bins-poiss}
ratios.normalization <- data.frame(bin = as.character(1:11))
ratios.normalization$surface_area <- NA
for (i in 1:11) {
  bin <- bins[[as.character(i)]]
  ratios.normalization$surface_area[ratios.normalization$bin == as.character(i)] <- length(ratios.simulated$Ratio[ratios.simulated$Ratio >= bin[1] &
    ratios.simulated$Ratio < bin[2]]) / length(ratios.simulated$Ratio)
}
ratios.normalization$bin_left <- sapply(bins, function(x) x[1])
ratios.normalization$bin_right <- sapply(bins, function(x) x[2])
ratios.normalization$bin_size <- ratios.normalization$bin_right - ratios.normalization$bin_left
ratios.normalization$bin_name <- c("1/4.5", "1/4", "1/3.5", "1/3", "1/2.5", "1/2", "1 - 1/2", "1 - 1/3", "1 - 1/3.5", "1 - 1/4", "1 - 1/4.5")
ratios.normalization$bin_alternative_name <- c("", "1:3", "", "1:2", "", "1:1", "", "2:1", "", "3:1", "")
```

## Counting the number of empirical observations in each bin and normalize

Below, we count the number of empirical observations in each bin and normalize them by dividing by the _bin width_.

```{r count-and-normalize-drumming-poiss}
# Code off- and on-isochronous
ratios$On_off_isoc <- NA
ratios$On_off_isoc <- factor(ratios$On_off_isoc, levels = c("Off", "On", "Other"))
ratios$On_off_isoc <- "Other"
ratios$On_off_isoc[ratios$Bin %in% c(5, 7)] <- "Off"
ratios$On_off_isoc[ratios$Bin == 6] <- "On"

# Count
ratios.isoc.counts.poiss <- dplyr::summarize(dplyr::group_by(ratios, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)

# Normalize
ratios.isoc.counts.poiss$On_isochronous_norm <- ratios.isoc.counts.poiss$On_isochronous / (ratios.normalization$bin_size[6] * ratios.isoc.counts$Total_obs)
ratios.isoc.counts.poiss$Off_isochronous_norm <- ratios.isoc.counts.poiss$Off_isochronous /
  ((ratios.normalization$bin_size[5] + ratios.normalization$bin_size[7]) * ratios.isoc.counts$Total_obs)
ratios.isoc.counts.poiss$Other_norm <- ratios.isoc.counts.poiss$Other / ((1 - sum(ratios.normalization$bin_size[5:7])) * ratios.isoc.counts$Total_obs)
```

The values in the normalized columns represent how much more above chance level the observed counts are. For example, a value of 1.0 means that the observed
counts are as expected by chance, a value of 2.0 means that the observed counts are twice what was expected by chance, and so on.

## Statistics

### Is the drumming random? (drumming ratios vs. random distribution)

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when IOIs are exponentially distributed (i.e. resulting from a Poisson point-process). We use the Kolmogorov-Smirnov test for this purpose.

#### All observations (both East and West subspecies) vs. Poisson-assumed

```{r ks-test-poiss-all}
ks.all.poiss <- ks.test(ratios$Ratio, "punif")
ks.all.poiss
```

*Across subspecies, the drumming ratios are not uniformly randomly distributed.*

#### Eastern chimps vs. Poisson-assumed

```{r ks-test-eac-poiss}
ks.eac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], "punif")
ks.eac.poiss
```

*Eastern chimps' drumming ratios are not uniformly randomly distributed.*

#### Western chimps vs. Poisson-assumed

```{r ks-test-wac-poiss}
ks.wac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Western"], "punif")
ks.wac.poiss
```

*Western chimps' drumming ratios are not uniformly randomly distributed.*

#### Are the distributions of Eastern and Western different from each other?

This of course does not depend on the null hypothesis of how the IOIs are distributed but is included here for completeness (as also done in the previous section).

```{r ks-test-eac-vs-wac-poiss}
ks.eac.vs.wac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios$Ratio[ratios$Subspecies == "Western"])
ks.eac.vs.wac.poiss
```

*The distributions of Eastern and Western are different from each other.*

### More on-isochronous vs. off-isochronous?

Here we use the the classic method of comparing the number of on-isochronous vs. off-isochronous ratios,
using the (until now common method for) normalization such that we divide by the _bin width_
(which assumes exponentially distributed IOIs resulting from a Poissant point-process).

#### All observations

```{r isoc-counts-wilcoxon-all-poiss}
# Paired Wilcoxon test
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm, ratios.isoc.counts.poiss$Off_isochronous_norm, paired = TRUE)
```

*Using the standard normalization procedure (i.e. dividing by bin width) we find that there are more on-isochronous ratios than off-isochronous ratios across subspecies.*

#### Eastern

```{r isoc-counts-wilcoxon-eac-poiss}
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"],
  ratios.isoc.counts.poiss$Off_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"],
  paired = TRUE
)
```

*Using the standard normalization procedure (i.e. dividing by bin width) we find that there are not more on-isochronous ratios than off-isochronous ratios in Eastern chimps.*

#### Western

```{r isoc-counts-wilcoxon-wac-poiss}
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  ratios.isoc.counts.poiss$Off_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  paired = TRUE
)
```

*Using the standard normalization procedure (i.e. dividing by bin width) we find that there are not more on-isochronous ratios than off-isochronous ratios in Western chimps.*

#### Are there more on-isochronous ratios in Western than in Eastern?

```{r isoc-counts-wilcoxon-eac-vs-wac-poiss}
wilcox.test(
  ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"]
)
print(paste("Median for East:", median(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"])))
print(paste("Median for West:", median(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"])))
```

*Using the standard normalization procedure (i.e. dividing by bin width) we find that there are more on-isochronous ratios in Western chimpanzees than in Eastern chimpanzees.*

# References

<div id="refs"></div>