---
title: "Drumming rhythm and tempo statistics"
author: Vesta Eleuteri & Jelle van der Werff
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    cards: true
    code_folding: hide
bibliography: references.bib
csl: apa.csl
---

# Introduction

This document accompanies the manuscript "Chimpanzee drumming shows rhythmicity and subspecies variation" @vestaeleuteriChimpanzeeDrummingShows2025.


# Preliminaries

## Load packages and data

```{r load-packages, echo = FALSE, warning = FALSE}
# for easy loading of packages
if (!require("pacman")) install.packages("pacman")

# load packages
p_load("here", "ggplot2", "dplyr", "nlme", "dplyr", "plyr", "knitr", "rmdformats", "kableExtra", "Matching", "kldtools", "effectsize", "lme4", "lmerTest", "psych")

here <- here::here

# for relative file paths (e.g. here('Input', 'file.csv'))
here::i_am(file.path("rhythm_stats.Rmd"))

# load data
ratios <- read.csv(here("dataframes", "drumming_ratios.csv"))
ibis <- read.csv(here("dataframes", "drumming_long.csv"))

# load custom ggplot theme
source(here("theme_drumming.R"))

# create simulated data based on uniform distribution
set.seed(1)
ratios.simulated <- runif(n = 1000001, min = min(ibis$IBI_ms), max = max(ibis$IBI_ms)) # IOIs

# calculate ratios
ratios.simulated <- ratios.simulated / (ratios.simulated + lead(ratios.simulated))
ratios.simulated <- ratios.simulated[!is.na(ratios.simulated)]
ratios.simulated <- as.data.frame(list(Ratio = ratios.simulated))

# set knitr options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, out.width = "100%", fig.height = 10, fig.width = 10, comment = NA)
```

## Defining bins and coding ratios categories

```{r bins}
bins <- list(
  "1" = c(0, 1 / 4.25), # 1 / 4.5
  "2" = c(1 / 4.25, 1 / 3.75), # 1 / 4 > 1:3
  "3" = c(1 / 3.75, 1 / 3.25), # 1 / 3.5
  "4" = c(1 / 3.25, 1 / 2.75), # 1 / 3 > 1:2
  "5" = c(1 / 2.75, 1 / 2.25), # 1 / 2.5
  "6" = c(1 / 2.25, 1 - 1 / 2.25), # 1 / 2 > 1:1
  "7" = c(1 - 1 / 2.25, 1 - 1 / 2.75), # 1 - 1 / 2.5
  "8" = c(1 - 1 / 2.75, 1 - 1 / 3.25), # 1 - 1 / 3 > 2:1
  "9" = c(1 - 1 / 3.25, 1 - 1 / 3.75), # 1 - 1 / 3.5
  "10" = c(1 - 1 / 3.75, 1 - 1 / 4.25), # 1 - 1 / 4 > 3:1
  "11" = c(1 - 1 / 4.25, 1) # 1 - 1 / 4.5
)

breaks <- c(0, 1 / 4.25, 1 / 3.75, 1 / 3.25, 1 / 2.75, 1 / 2.25, 1 - 1 / 2.25, 1 - 1 / 2.75, 1 - 1 / 3.25, 1 - 1 / 3.75, 1 - 1 / 4.25, 1)

labels <- names(bins)

ratios$Bin <- cut(ratios$Ratio, breaks = breaks, labels = labels)
ratios.simulated$Bin <- cut(ratios.simulated$Ratio, breaks = breaks, labels = labels)
```

## Reordering/renaming variables etc.

```{r reorder}
# Change levels order
ibis$Subspecies <- factor(ibis$Subspecies, levels = c("Western", "Eastern"))
ibis$Population <- factor(ibis$Population, levels = c("Fongoli", "Nimba", "Taï", "Gombe", "Kibale", "Budongo"))
ibis$Community <- factor(ibis$Community, levels = c(
  "Fongoli", "Tongbongbon", "Gahtoy", "East", "Middle", "North", "South",
  "Kasekela", "Kanyawara", "Sonso", "Waibira"
))

ratios$Subspecies <- factor(ratios$Subspecies, levels = c("Western", "Eastern"))
ratios$Population <- factor(ratios$Population, levels = c("Fongoli", "Nimba", "Taï", "Gombe", "Kibale", "Budongo"))
ratios$Community <- factor(ratios$Community, levels = c(
  "Fongoli", "Tongbongbon", "Gahtoy", "East", "Middle", "North", "South",
  "Kasekela", "Kanyawara", "Sonso", "Waibira"
))
```

# Descriptives (inter-hit intervals and ratios)

```{r descriptives}
subspecies.table <- ddply(ibis, .(Subspecies), summarize,
  mean = mean(IBI_ms),
  median = median(IBI_ms),
  sd = sd(IBI_ms),
  min = min(IBI_ms),
  max = max(IBI_ms)
)
kable(subspecies.table, format = "html", caption = "Descriptives of inter-hit intervals (ms) by subspecies") %>%
  kable_styling("striped", full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:6, width = "100px")

overall.table <- as.data.frame(ddply(ibis, .(), summarize,
  mean = mean(IBI_ms),
  median = median(IBI_ms),
  sd = sd(IBI_ms),
  min = min(IBI_ms),
  max = max(IBI_ms)
))
overall.table <- overall.table[, 2:6]

kable(overall.table, format = "html", caption = "Descriptives of inter-hit intervals (ms) overall") %>%
  kable_styling("striped", full_width = FALSE)
```


# Main text: Analyses for non-random timing, isochrony, and shorter-longer adjacent inter-hit intervals in drumming

## Introduction

Our main analysis relies on the method devised by @roeskeCategoricalRhythmsAre2020, in which the relationships between adjacent inter-onset/inter-hit intervals
(IOIs/IHIs) are described by the formula: $r_k = \frac{IOI_k}{IOI_k + IOI_{k+1}}$. This formula results in ratios $r_k$ that range from 0 to 1, where e.g. 0.5
indicates isochrony.

Recently, @jadoulHiddenAssumptionsInteger2025 have highlighted a number of issues with the usual method of doing statistics with these ratios, and have improved the
method as hitherto used. Our analysis takes into account these new insights.

Classically, the null hypothesis is that observed IOIs are distributed according to a random uniform distribution. However, the statistical tests that are
generally used do not conform to this null hypothesis [@jadoulHiddenAssumptionsInteger2025]. According to @jadoulHiddenAssumptionsInteger2025, the assumption of a uniformly random
distribution of IOIs as the null-hypothesis means that the ratios are distributed according to the distribution plotted below. The exact shape of this
distribution is dependent on the left and right bounds of the random uniform distribution used for sampling the IOIs. However, the shape will remain similar
to the one plotted below.

As left (a) and right (b) bounds of the uniform distribution of IOIs, in our analysis we use the minimum observed IOI duration and the maximum observed IOI
duration, respectively.

The dotted lines in the plot are the histogram bins, following @roeskeCategoricalRhythmsAre2020, which represent the integer ratios 1:3 (bin centered around 0.25),
1:2 (bin centered around 0.33), 1:1 (isochrony; bin centered around 0.5), 2:1 (bin centered around 0.66), and 3:1 (bin centered around 0.75).
The bins in between these integer ratios are the off-integer bins.

```{r uniform-random-distribution}
ggplot(ratios.simulated, aes(x = Ratio)) +
  geom_density(linewidth = 2) +
  theme_drumming() +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.5, colour = "grey") +
  labs(x = "Ratio", y = "Probability density", title = "Random ratios distribution", subtitle = "Assumes uniformly random IOIs") +
  scale_x_continuous(
    breaks = c(0, 0.25, 0.3333, 0.5, 0.6666, 0.75, 1), limits = c(0, 1),
    labels = c("0.0", "0.25\n1:3", "0.33\n1:2", "0.5\n1:1", "0.66\n3:2", "0.75\n3:1", "1.0")
  )
```

We define the bins following Roeske et al. (2020):

```{r surface-areas}
ratios.bins <- data.frame(bin = as.character(1:11))

ratios.bins$bin_left <- sapply(bins, function(x) x[1])
ratios.bins$bin_right <- sapply(bins, function(x) x[2])
ratios.bins$bin_name <- c("", "1/4", "1/3.5", "1/3", "1/2.5", "1/2", "1 - 1/2.5", "1 - 1/3", "1 - 1/3.5", "1 - 1/4", "")
ratios.bins$bin_alternative_name <- c("", "1:3", "", "1:2", "", "1:1", "", "2:1", "", "3:1", "")

kable(ratios.bins,
  format = "html", caption = "Bins definition", col.names =
    c("Bin", "Bin (left bound)", "Bin (right bound)", "Bin center", "Integer ratio")
) %>%
  kable_styling("striped", full_width = FALSE)
```

Then, we count the number of observations in each bin.

```{r count-and-normalize-drumming}
# Code off- and on-isochronous
ratios$On_off_isoc <- NA
ratios$On_off_isoc <- factor(ratios$On_off_isoc, levels = c("Off", "On", "Other"))
ratios$On_off_isoc <- "Other"
ratios$On_off_isoc[ratios$Bin %in% c(5, 7)] <- "Off"
ratios$On_off_isoc[ratios$Bin == 6] <- "On"

# Count
ratios.isoc.counts <- dplyr::summarize(dplyr::group_by(ratios, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)

kable(head(ratios.isoc.counts, 10),
  format = "html", align = "l",
  caption = "Counts of on- and off-isochronous drumming ratios (first 10 rows)",
  col.names = c(
    "Subspecies", "Population", "Community", "Community (Individual)", "On-isochronous count", "Off-isochronous count", "Other ratios count",
    "Total n ratios observed"
  )
) %>%
  kable_styling("striped", full_width = FALSE)
```

## Plot of ratios distributions by subspecies

```{r plot-ratios-distribution}
labels <- as.data.frame(list(
  x = c(0.25, 0.335, 0.5, 0.665, 0.75),
  y = c(0.1, 0.1, 0.1, 0.1, 0.1),
  label = c("1:3", "1:2", "1:1", "2:1", "3:1"),
  size = c(4, 6, 10, 6, 4)
))

ggplot(ratios, aes(x = Ratio, fill = Subspecies)) +
  geom_density(linewidth = 0, alpha = 0.7) +
  geom_density(data = ratios.simulated, aes(x = Ratio), linewidth = 2, colour = "yellow", inherit.aes = FALSE, show.legend = FALSE) +
  theme_drumming() +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.5, colour = "darkgrey") +
  labs(y = "Density", x = "r", fill = "", title = "B - Subspecies rhythmic patterns") +
  geom_label(data = labels, aes(x = x, y = y, label = label), size = 10, family = "Helvetica", inherit.aes = FALSE) +
  theme(
    plot.title = element_text(size = 38),
    axis.title.x = element_text(size = 32),
    axis.title.y = element_text(size = 32),
    axis.text.x = element_text(size = 28),
    axis.text.y = element_text(size = 28),
    legend.title = element_text(size = 32),
    legend.text = element_text(size = 32)
  ) +
  theme(plot.margin = margin(1, 0, 1, 1, "cm")) +
  guides(fill = "none")

ggsave(here("plots", "integer_ratios_vs_random.pdf"), width = 15, height = 10)
ggsave(here("plots", "integer_ratios_vs_random.png"), width = 15, height = 10)
ggsave(here("plots", "integer_ratios_vs_random.svg"), width = 15, height = 10)
```


## Statistics

Here we test:

1. Whether the drumming ratios are randomly distributed (i.e. whether the distribution of observed ratios is different from a distribution of ratios derived from uniformly random inter-hit intervals).
2. Whether the distributions of eastern and western differ from each other.
3. Whether there are more on-isochronous ratios than expected by chance.
4. Whether there are more on-isochronous ratios in eastern vs. western chimpanzees.
5. Whether there is more alternation of shorter-longer inter-hit intervals in eastern vs. western chimpanzees.


### Do chimpanzees show non-random timing in drumming? 

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when inter-hit intervals (IHIs) are uniformly randomly distributed (which we simulated above). We use Kolmogorov-Smirnov tests for this purpose.

#### All observations (both eastern and western chimpanzee drumming) vs. uniformly random

```{r ks-test}
ks.all <- ks.test(ratios$Ratio, ratios.simulated$Ratio, alternative = "two.sided")
ks.all
```

**Across subspecies, for all chimpanzees the drumming is not random.**

#### Western chimpanzee vs. uniformly random

```{r ks-test-wac}
ks.wac <- ks.test(ratios$Ratio[ratios$Subspecies == "Western"], ratios.simulated$Ratio, alternative = "two.sided")
ks.wac
```

**For western chimpanzees the drumming is not random.**

#### Eastern chimpanzees vs. uniformly random

```{r ks-test-eac}
ks.eac <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios.simulated$Ratio, alternative = "two.sided")
ks.eac
```

**For eastern chimpanzees the drumming is not random.**

### Do eastern and western chimpanzees show different distributions of ratios?

```{r ks-test-eac-vs-wac}
ks.eac.vs.wac <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios$Ratio[ratios$Subspecies == "Western"], alternative = "two.sided")
ks.eac.vs.wac
```

**Eastern and western chimpanzees show different distributions of ratios.** 

### Do chimpanzees show isochrony in drumming?

Here we compare the proportion of on-isochronous ratios that were observed to the number of on-isochronous ratios that can be expected by chance (i.e., when assuming uniformly random IHIs).

We convert the number of on-isochronous observations to probabilities using: $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$. We do this for each individual chimpanzee separately. The individuals' probabilities are then compared to chance probability using Wilcoxon signed-rank tests. Chance probability was calculated using the formula provided by @jadoulHiddenAssumptionsInteger2025:

$$\frac{1}{2} - \frac{(4b - 5a)^2}{40 (b - a)^2}$$

This yielded a value of 0.2042091.

```{r isoc-vs-random-preparation}
# Calculate the surface area of the isochronous bin using the formula from Jadoul et al. (2025)
a <- min(ibis$IBI_ms)
b <- max(ibis$IBI_ms)
surface_area_isochronous <- ((1 / 2) - (((4 * b - 5 * a)^2) / (40 * (b - a)^2)))

# Multiply by two (there's one bin on each side of 0.5 when using this formula)
surface_area_isochronous <- surface_area_isochronous * 2 # 0.2042091

# Subtract surface area of isochronous bin from empirical probabilities
ratios.isoc.counts$Isoc_prob <- ratios.isoc.counts$On_isochronous / ratios.isoc.counts$Total_obs
ratios.isoc.counts$Isoc_prob_normalized <- ratios.isoc.counts$Isoc_prob - surface_area_isochronous
```

```{r isoc-vs-random-plot}
set.seed(2)
ggplot(ratios.isoc.counts, aes(y = Isoc_prob * 100, x = Subspecies, fill = Subspecies)) +
  geom_boxplot(colour = "black", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, show.legend = FALSE) +
  theme_drumming() +
  labs(
    y = "", x = "",
  title = "") +
  geom_hline(yintercept = surface_area_isochronous * 100, linetype = "dashed", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

ggsave(here("plots", "isochronous_ratios_vs_random.pdf"), width = 7, height = 5)
ggsave(here("plots", "isochronous_ratios_vs_random.png"), width = 7, height = 5)
ggsave(here("plots", "isochronous_ratios_vs_random.svg"), width = 7, height = 5)

```

#### All observations (both eastern and western chimpanzee drumming)

```{r isoc-vs-random-wilcoxon-all}
isoc.vs.random.wsrt.all <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized, alternative = "two.sided")
isoc.vs.random.wsrt.all
print(paste("Median for all chimps:", median(ratios.isoc.counts$Isoc_prob_normalized)))
```

**Across subspecies, there are not more on-isochronous ratios than expected by chance.**

#### Western chimpanzees

```{r isoc-vs-random-wilcoxon-wac}
isoc.vs.random.wsrt.wac <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Western"],
  alternative = "two.sided"
) # Significant
isoc.vs.random.wsrt.wac
print(paste("Median for Western:", median(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Western"])))
```

**In Western chimpanzees, there are significantly _more_ on-isochronous ratios than expected by chance.**

#### Eastern chimpanzees

```{r isoc-vs-random-wilcoxon-eac}
isoc.vs.random.wsrt.eac <- wilcox.test(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Eastern"],
  alternative = "two.sided"
) # Significant
isoc.vs.random.wsrt.eac
print(paste("Median for Eastern:", median(ratios.isoc.counts$Isoc_prob_normalized[ratios.isoc.counts$Subspecies == "Eastern"])))
```

**In Eastern chimpanzees, there are significantly _fewer_ on-isochronous ratios than expected by chance.**

### Is there more isochrony in eastern vs. western chimpanzees?

Here we compare the empirical probabilities (i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) of on-isochronous
ratios between eastern and western chimpanzees using a Wilcoxon rank-sum test.

```{r isoc-counts-wilcoxon-eac-vs-wac}
wilcox.test(
  ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Western"],
  ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Eastern"]
)
print(paste("Median for East:", median(ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Eastern"])))
print(paste("Median for West:", median(ratios.isoc.counts$Isoc_prob[ratios.isoc.counts$Subspecies == "Western"])))
```

**There are more on-isochronous ratios in western than in eastern chimpanzees.**

### Do eastern chimpanzees alternate shorter-longer inter-hit intervals more than western chimpanzees? 

We first extract the lag-one autocorrelations values by subspecies and drumming bout

```{r get-autocorrelations}
acf.cv <- ddply(ibis, .(Subspecies, Population, Community, Individual, Drumming_bout, N_hits), summarize,
    lag_one = acf(IBI, lag.max = 1, plot = FALSE)$acf[2],
    cv = sd(IBI) / mean(IBI)
)

acf.cv.bysubspecies <- ddply(acf.cv, .(Subspecies), summarize,
    mean_lag_one = mean(lag_one, na.rm = TRUE),
    sd_lag_one = sd(lag_one, na.rm = TRUE),
    mean_cv = mean(cv, na.rm = TRUE),
    sd_cv = sd(cv, na.rm = TRUE)
)

kable(acf.cv.bysubspecies, format = "html", escape = FALSE) %>%
    kable_styling("striped", full_width = FALSE)
```

We then plot the lag-one autocorrelations 

```{r plot-autocorrelations}
ggplot(acf.cv, aes(y = Subspecies, x = lag_one, colour = Subspecies)) +
    stat_summary(geom = "pointrange", show.legend = FALSE) +
    theme_drumming() +
    labs(y = "", x = "") +
    coord_cartesian(xlim = c(-0.5, 0)) +
    scale_x_continuous(breaks = c(-0.5, -0.25, 0)) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Make panel background transparent
        plot.background = element_rect(fill = "transparent", color = NA), # Make plot background transparent
        panel.grid.major = element_blank(), # Remove major grid lines
        panel.grid.minor = element_blank(), # Remove minor grid lines
        panel.grid.major.x = element_line(color = "gray80", size = 0.5), # Add vertical major grid lines
        panel.grid.major.y = element_blank(), # Remove major horizontal grid lines
        legend.background = element_rect(fill = "transparent", color = NA), # Make legend background transparent
        legend.box.background = element_rect(fill = "transparent", color = NA) # Make legend box background transparent
    )
```

Then we compare the lag-one autocorrelations by subspecies using a t-test

```{r do-t-test}
# Check equality of variances
acf.east <- acf.cv$lag_one[acf.cv$Subspecies == "Eastern"]
acf.west <- acf.cv$lag_one[acf.cv$Subspecies == "Western"]
var.test(acf.east, acf.west)

# Run t-test
acf.t.test <- t.test(acf.east, acf.west)

# ACF by subspecies
acf.by.subspecies <- as.data.frame(list(
    test_statistic = acf.t.test$statistic,
    p_value = acf.t.test$p.value,
    cohens_d = mean(acf.east, na.rm = TRUE) - mean(acf.west, na.rm = TRUE) / sqrt((var(acf.east, na.rm = TRUE) + var(acf.west, na.rm = TRUE)) / 2),
    df = acf.t.test$parameter,
    method = acf.t.test$method
))

kable(acf.by.subspecies, format = "html", escape = FALSE) %>%
    kable_styling("striped", full_width = FALSE)
```

**Eastern chimpanzees alternate shorter and longer inter-hit intervals in drumming more often than western chimpanzees.**


# Main text: Analysis for subspecies variation in drumming tempo

## Do western chimpanzees drum with a faster tempo than eastern chimpanzees?

We first extract the drumming tempo per drumming bout and make a drumming tempo plot

```{r boxplot}
tempi <- ddply(ibis, .(Subspecies, Population, Community, Individual, Drumming_bout), summarize,
  tempo = mean(IBI_ms)
)

ggplot(tempi, aes(x = Subspecies, y = tempo, fill = Subspecies)) +
  geom_boxplot(alpha = 0.7) +
  geom_point(show.legend = FALSE, alpha = 0.7, position = position_jitter(width=0.1)) +
  labs(y = "Tempo [ms]", title = "Subspecies drumming tempo") +
  theme_drumming() +
  theme(
    plot.title = element_text(size = 40),
    axis.title.x = element_text(size = 34),
    axis.title.y = element_text(size = 34),
    axis.text.x = element_text(size = 30),
    axis.text.y = element_text(size = 30),
    legend.title = element_text(size = 34, hjust=1),
    legend.text = element_text(size = 34, hjust=1),
    legend.position = "none",  
    legend.key.size = unit(1, "cm"), 
    legend.spacing.y = unit(0.5, "cm")) 
```

We then compare drumming tempo by subspecies using a glmm

```{r mixed-model}
model <- lmer(tempo ~ Subspecies + (1 | Individual), data = tempi)
summary(model)
effectsize(model)
```

```{r model-comparison}
tempo.null.model <- lmer(tempo ~ (1 | Individual), data = tempi, REML = FALSE)
tempo.comparison <- as.data.frame(anova(tempo.null.model, model, test = "Chisq"))
tempo.comparison.chisquare <- round(tempo.comparison, 3)
tempo.comparison.chisquare
```

**Western chimpanzees drum with a faster tempo than eastern chimpanzees.**


# Supplement: Analyses for non-random timing, isochrony, and shorter-longer adjacent inter-hit intervals in >= 5 hit drumming bouts.

## Preparations

We subset the data to only include drumming bouts with 5 or more hits.

```{r copy-df-and-filter-5hits}
ratios.5plushits <- ratios
ratios.5plushits <- ratios.5plushits %>%
  group_by(Drumming_bout) %>%
  filter(n() >= 3) %>% # remove drums < 3 ratios.5plushits (4 IBIs/5 beats)
  group_by(Com_indiv_code) %>% # remove individuals with <3 drums
  filter(n_distinct(Drumming_bout) >= 3) %>%
  group_by(Community) %>% # remove communities with <1 individual
  filter(n_distinct(Individual) > 1) %>%
  ungroup()
```

Below, we count the number of empirical observations in each bin for drumming bouts with 5 hits or more.

```{r count-and-normalize-drumming-5hits}
# Code off- and on-isochronous
ratios.5plushits$On_off_isoc <- NA
ratios.5plushits$On_off_isoc <- factor(ratios.5plushits$On_off_isoc, levels = c("Off", "On", "Other"))
ratios.5plushits$On_off_isoc <- "Other"
ratios.5plushits$On_off_isoc[ratios.5plushits$Bin %in% c(5, 7)] <- "Off"
ratios.5plushits$On_off_isoc[ratios.5plushits$Bin == 6] <- "On"

# Count
ratios.5plushits.isoc.counts <- dplyr::summarize(dplyr::group_by(ratios.5plushits, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)
```

### Plot of ratios distributions by subspecies for 5 or more hit drumming bouts

```{r plot-ratios-distribution-5hits}
labels <- as.data.frame(list(
  x = c(0.25, 0.335, 0.5, 0.665, 0.75),
  y = c(0.1, 0.1, 0.1, 0.1, 0.1),
  label = c("1:3", "1:2", "1:1", "2:1", "3:1"),
  size = c(4, 6, 10, 6, 4)
))

ggplot(ratios.5plushits, aes(x = Ratio, fill = Subspecies)) +
  geom_density(linewidth = 0, alpha = 0.7) +
  geom_density(data = ratios.simulated, aes(x = Ratio), linewidth = 2, colour = "yellow", inherit.aes = FALSE, show.legend = FALSE) +
  theme_drumming() +
  geom_vline(xintercept = breaks, linetype = "dashed", linewidth = 0.5, colour = "grey") +
  labs(
    y = "Density", x = "r", fill = "Subspecies", title = "Subspecies rhythmic patterns for drumming bouts with at least 5 hits",
  ) +
  geom_label(data = labels, aes(x = x, y = y, label = label), size = 10, family = "Helvetica", inherit.aes = FALSE) +
  theme(
    plot.title = element_text(size = 40),
    axis.title.x = element_text(size = 34),
    axis.title.y = element_text(size = 34),
    axis.text.x = element_text(size = 30),
    axis.text.y = element_text(size = 30),
    legend.title = element_text(size = 34, hjust = 1),
    legend.text = element_text(size = 34, hjust = 1),
    legend.position = "right",
    legend.key.size = unit(1, "cm"),
    legend.spacing.y = unit(0.5, "cm")
  )

ggsave(here("plots", "integer_ratios_vs_random_5hitsormore.pdf"), width = 20, height = 15)
```

## Statistics

### Do chimpanzees show non-random timing in drumming bouts with >=5 hits? (drumming ratios vs. random distribution)

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when inter-hit intervals (IHIs) are uniformly randomly distributed. We use the Kolmogorov-Smirnov test for this purpose.

#### All observations (both eastern and western chimpanzee drumming) vs. uniformly random

```{r ks-test-5hits}
ks.all <- ks.test(ratios.5plushits$Ratio, ratios.simulated$Ratio, alternative = "two.sided")
ks.all
```

**Across subspecies, for all chimpanzees the drumming bouts with 5 or more hits are not random.**

#### Eastern chimpanzees vs. uniformly random

```{r ks-test-eac-5hits}
ks.eac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Eastern"], ratios.simulated$Ratio, alternative = "two.sided")
ks.eac
```

**For eastern chimpanzees, the drumming bouts with 5 or more hits are random.**

#### Western chimpanzees vs. uniformly random

```{r ks-test-wac-5hits}
ks.wac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Western"], ratios.simulated$Ratio, alternative = "two.sided")
ks.wac
```

**For western chimpanzees, the drumming bouts with 5 or more hits are not random.**

### Do eastern and western chimpanzees show different distributions of ratios?

```{r ks-test-eac-vs-wac-5hits}
ks.eac.vs.wac <- ks.test(ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Eastern"],
  ratios.5plushits$Ratio[ratios.5plushits$Subspecies == "Western"],
  alternative = "two.sided"
)
ks.eac.vs.wac
```

**Eastern and western chimpanzees show different distributions of ratios in drumming bouts with 5 or more hits.** 

### Are there more on-isochronous ratios than expected by chance?

Here we compare the proportion of on-isochronous ratios that were observed to the number of on-isochronous ratios that can be expected by chance (i.e., when assuming uniformly random IHIs).

We convert the number of on-isochronous observations to probabilities using: $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$. We do this for each individual chimpanzee separately. The individuals' probabilities are then compared to chance probability using Wilcoxon signed-rank tests. Chance probability was calculated using the formula provided by @jadoulHiddenAssumptionsInteger2025:

$$\frac{1}{2} - \frac{(4b - 5a)^2}{40 (b - a)^2}$$

```{r isoc-vs-random-preparation-5hits}
# Calculate the surface area of the isochronous bin using the formula from Jadoul et al. (2025)
a <- min(ibis$IBI_ms)
b <- max(ibis$IBI_ms)
surface_area_isochronous <- ((1 / 2) - (((4 * b - 5 * a)^2) / (40 * (b - a)^2)))

# Multiply by two (there's one bin on each side of 0.5 when using this formula)
surface_area_isochronous <- surface_area_isochronous * 2 # 0.2042091

# Subtract surface area of isochronous bin from empirical probabilities
ratios.5plushits.isoc.counts$Isoc_prob <- ratios.5plushits.isoc.counts$On_isochronous / ratios.5plushits.isoc.counts$Total_obs
ratios.5plushits.isoc.counts$Isoc_prob_normalized <- ratios.5plushits.isoc.counts$Isoc_prob - surface_area_isochronous
```

```{r isoc-vs-random-plot-5hits}
ggplot(ratios.5plushits.isoc.counts, aes(y = Isoc_prob * 100, x = Subspecies, fill = Subspecies)) +
  geom_boxplot(colour = "black", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, show.legend = FALSE) +
  theme_drumming() +
  labs(
    y = "Percentage of isochronous drumming ratios per individual", x = "Subspecies",
    title = "Percentage of isochronous ratios compared to chance level",
    subtitle = "Chance level indicated by dotted line is based on uniformly randomly distributed IOIs.\nOnly drumming bouts with 5 or more hits are included.",
  ) +
  geom_hline(yintercept = surface_area_isochronous * 100, linetype = "dashed", linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))
```

#### All observations (both eastern and western chimpanzee drumming)

```{r isoc-vs-random-wilcoxon-all-5hits}
isoc.vs.random.wsrt.all <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized, alternative = "two.sided")
isoc.vs.random.wsrt.all
print(paste("Median for all chimps:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized)))
```

**Across subspecies, for all chimpanzees there are not more on-isochronous ratios than expected by chance in drumming bouts with 5 or more hits.**

#### Eastern chimpanzees

```{r isoc-vs-random-wilcoxon-eac-5hits}
isoc.vs.random.wsrt.eac <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Eastern"], alternative = "two.sided") # Significant
isoc.vs.random.wsrt.eac
print(paste("Median for Eastern:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Eastern"])))
```

**In eastern chimpanzees, there are not  more or fewer on-isochronous ratios than expected by chance in drumming bouts with 5 or more hits.**

#### Western chimpanzees

```{r isoc-vs-random-wilcoxon-wac-5hits}
isoc.vs.random.wsrt.wac <- wilcox.test(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Western"], alternative = "two.sided") # Significant
isoc.vs.random.wsrt.wac
print(paste("Median for Western:", median(ratios.5plushits.isoc.counts$Isoc_prob_normalized[ratios.5plushits.isoc.counts$Subspecies == "Western"])))
```

**In western chimpanzees, there are significantly _more_ on-isochronous ratios than expected by chance in drumming bouts with 5 or more hits.**

### Is there more isochrony in eastern vs. western chimpanzees?

Here we compare the empirical probabilities (i.e. $\frac{\textrm{number of isochronous observations}}{\textrm{total number of observations}}$) of on-isochronous
ratios between eastern and western chimpanzees using a Wilcoxon rank-sum test.

```{r isoc-counts-wilcoxon-eac-vs-wac-5hits}
wilcox.test(
  ratios.5plushits.isoc.counts$Isoc_prob[ratios.5plushits.isoc.counts$Subspecies == "Western"],
  ratios.5plushits.isoc.counts$Isoc_prob[ratios.5plushits.isoc.counts$Subspecies == "Eastern"]
)
```

**There are more on-isochronous ratios in western than in eastern chimpanzees in drumming bouts with 5 or more hits.**

### Do eastern chimpanzees alternate shorter-longer inter-hit intervals more than western chimpanzees? 

We subset the data to only include drumming bouts with 5 or more hits.

```{r filter-5plushits}
ibis.5plushits <- ibis
ibis.5plushits <- ibis.5plushits %>%
  group_by(Drumming_bout) %>%
  filter(n() >= 4) %>% # remove drums < 4 IHIs
  group_by(Com_indiv_code) %>% # remove individuals with <3 drums
  filter(n_distinct(Drumming_bout) >= 3) %>%
  group_by(Community) %>% # remove communities with <1 individual
  filter(n_distinct(Individual) > 1) %>%
  ungroup()

```

We first extract the lag-one autocorrelations values by subspecies and drumming bout

```{r get-autocorrelations-5plushits}
acf.cv.5plushits <- ddply(ibis.5plushits, .(Subspecies, Population, Community, Individual, Drumming_bout, N_hits), summarize,
    lag_one = acf(IBI, lag.max = 1, plot = FALSE)$acf[2],
    cv = sd(IBI) / mean(IBI)
)

acf.cv.bysubspecies.5plushits <- ddply(acf.cv.5plushits, .(Subspecies), summarize,
    mean_lag_one = mean(lag_one, na.rm = TRUE),
    sd_lag_one = sd(lag_one, na.rm = TRUE),
    mean_cv = mean(cv, na.rm = TRUE),
    sd_cv = sd(cv, na.rm = TRUE)
)

kable(acf.cv.bysubspecies.5plushits, format = "html", escape = FALSE) %>%
    kable_styling("striped", full_width = FALSE)
```

We then plot the lag-one autocorrelations

```{r plot-autocorrelations-5plushits}
ggplot(acf.cv.5plushits, aes(y = Subspecies, x = lag_one, colour = Subspecies)) +
    stat_summary(geom = "pointrange", show.legend = FALSE) +
    theme_drumming() +
    labs(y = "", x = "") +
    coord_cartesian(xlim = c(-0.5, 0)) +
    scale_x_continuous(breaks = c(-0.5, -0.25, 0)) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Make panel background transparent
        plot.background = element_rect(fill = "transparent", color = NA), # Make plot background transparent
        panel.grid.major = element_blank(), # Remove major grid lines
        panel.grid.minor = element_blank(), # Remove minor grid lines
        panel.grid.major.x = element_line(color = "gray80", size = 0.5), # Add vertical major grid lines
        panel.grid.major.y = element_blank(), # Remove major horizontal grid lines
        legend.background = element_rect(fill = "transparent", color = NA), # Make legend background transparent
        legend.box.background = element_rect(fill = "transparent", color = NA) # Make legend box background transparent
    )
```

Then we compare the lag-one autocorrelations by subspecies using a t-test

```{r do-t-test-5plushits}
# Check equality of variances
acf.5plushits.east <- acf.cv.5plushits$lag_one[acf.cv.5plushits$Subspecies == "Eastern"]
acf.5plushits.west <- acf.cv.5plushits$lag_one[acf.cv.5plushits$Subspecies == "Western"]
var.test(acf.5plushits.east, acf.5plushits.west)

# Run t-test
acf.5plushits.t.test <- t.test(acf.5plushits.east, acf.5plushits.west)

# ACF by subspecies
acf.5plushits.by.subspecies <- as.data.frame(list(
    test_statistic = acf.5plushits.t.test$statistic,
    p_value = acf.5plushits.t.test$p.value,
    cohens_d = mean(acf.5plushits.east, na.rm = TRUE) - mean(acf.5plushits.west, na.rm = TRUE) / sqrt((var(acf.5plushits.east, na.rm = TRUE) +
        var(acf.5plushits.west, na.rm = TRUE)) / 2),
    df = acf.5plushits.t.test$parameter,
    method = acf.5plushits.t.test$method
))

kable(acf.5plushits.by.subspecies, format = "html", escape = FALSE) %>%
    kable_styling("striped", full_width = FALSE)
```

**The alternation of shorter and longer inter-hit intervals does not differ between eastern and western chimpanzees in drumming bouts with 5 or more hits.**


# Supplement: Analyses for non-random timing and isochrony fully replicating previous studies

## Introduction
We here use the method as has been used until now in the literature, which, due to the normalization procedure
(i.e. dividing by the bin _width_ rather than the bin _surface area_), assumes exponentially distributed IHIs/IOIs resulting from a Poisson point-process
[@jadoulHiddenAssumptionsInteger2025].

## Defining bins, bin widths, etc.

```{r bins-poiss}
ratios.normalization <- data.frame(bin = as.character(1:11))
ratios.normalization$surface_area <- NA
for (i in 1:11) {
  bin <- bins[[as.character(i)]]
  ratios.normalization$surface_area[ratios.normalization$bin == as.character(i)] <- length(ratios.simulated$Ratio[ratios.simulated$Ratio >= bin[1] &
    ratios.simulated$Ratio < bin[2]]) / length(ratios.simulated$Ratio)
}
ratios.normalization$bin_left <- sapply(bins, function(x) x[1])
ratios.normalization$bin_right <- sapply(bins, function(x) x[2])
ratios.normalization$bin_size <- ratios.normalization$bin_right - ratios.normalization$bin_left
ratios.normalization$bin_name <- c("1/4.5", "1/4", "1/3.5", "1/3", "1/2.5", "1/2", "1 - 1/2", "1 - 1/3", "1 - 1/3.5", "1 - 1/4", "1 - 1/4.5")
ratios.normalization$bin_alternative_name <- c("", "1:3", "", "1:2", "", "1:1", "", "2:1", "", "3:1", "")
```

## Counting the number of empirical observations in each bin and normalize

Below, we count the number of empirical observations in each bin and normalize them by dividing by the _bin width_.

```{r count-and-normalize-drumming-poiss}
# Code off- and on-isochronous
ratios$On_off_isoc <- NA
ratios$On_off_isoc <- factor(ratios$On_off_isoc, levels = c("Off", "On", "Other"))
ratios$On_off_isoc <- "Other"
ratios$On_off_isoc[ratios$Bin %in% c(5, 7)] <- "Off"
ratios$On_off_isoc[ratios$Bin == 6] <- "On"

# Count
ratios.isoc.counts.poiss <- dplyr::summarize(dplyr::group_by(ratios, Subspecies, Population, Community, Com_indiv_code),
  On_isochronous = sum(On_off_isoc == "On"), Off_isochronous = sum(On_off_isoc == "Off"), Other = sum(On_off_isoc == "Other"), Total_obs = n()
)

# Normalize
ratios.isoc.counts.poiss$On_isochronous_norm <- ratios.isoc.counts.poiss$On_isochronous / (ratios.normalization$bin_size[6] * ratios.isoc.counts$Total_obs)
ratios.isoc.counts.poiss$Off_isochronous_norm <- ratios.isoc.counts.poiss$Off_isochronous /
  ((ratios.normalization$bin_size[5] + ratios.normalization$bin_size[7]) * ratios.isoc.counts$Total_obs)
ratios.isoc.counts.poiss$Other_norm <- ratios.isoc.counts.poiss$Other / ((1 - sum(ratios.normalization$bin_size[5:7])) * ratios.isoc.counts$Total_obs)
```

The values in the normalized columns represent how much more above chance level the observed counts are. For example, a value of 1.0 means that the observed
counts are as expected by chance, a value of 2.0 means that the observed counts are twice what was expected by chance, and so on.

## Statistics

### Do chimpanzees drum with non-random timing (drumming ratios vs. random distribution)?

Here we answer the question: "Is the drumming random?" by comparing the distributions of observed ratios with the distribution of ratios as expected
when IHIs/IOIs are exponentially distributed (i.e. resulting from a Poisson point-process). We use the Kolmogorov-Smirnov test for this purpose.

#### All observations (both eastern and western chimpanzee drumming) vs. Poisson-assumed

```{r ks-test-poiss-all}
ks.all.poiss <- ks.test(ratios$Ratio, "punif")
ks.all.poiss
```

**Across subspecies, for all chimpanzees the drumming is not random.**

#### Eastern chimpanzees vs. Poisson-assumed

```{r ks-test-eac-poiss}
ks.eac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], "punif")
ks.eac.poiss
```

**For eastern chimpanzees, the drumming is not random.**

#### Western chimpanzees vs. Poisson-assumed

```{r ks-test-wac-poiss}
ks.wac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Western"], "punif")
ks.wac.poiss
```

**For western chimpanzees, the drumming is not random.**

#### Do eastern and western chimpanzees show different distributions of ratios?

This of course does not depend on the null hypothesis of how the IHIs are distributed but is included here for completeness (as also done in the previous section).

```{r ks-test-eac-vs-wac-poiss}
ks.eac.vs.wac.poiss <- ks.test(ratios$Ratio[ratios$Subspecies == "Eastern"], ratios$Ratio[ratios$Subspecies == "Western"])
ks.eac.vs.wac.poiss
```

**Eastern and western chimpanzees show different distributions of ratios.** 

### More on-isochronous vs. off-isochronous?

Here we use the the classic method of comparing the number of on-isochronous vs. off-isochronous ratios,
using the (until now common method for) normalization such that we divide by the _bin width_
(which assumes exponentially distributed IHIs/IOIs resulting from a Poissant point-process).

#### All observations (both eastern and western chimpanzee drumming)

```{r isoc-counts-wilcoxon-all-poiss}
# Paired Wilcoxon test
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm, ratios.isoc.counts.poiss$Off_isochronous_norm, paired = TRUE)
```

**Using the standard normalization procedure (i.e. dividing by bin width), we find that there are more on-isochronous ratios than off-isochronous ratios across subspecies.**

#### Eastern chimpanzees

```{r isoc-counts-wilcoxon-eac-poiss}
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"],
  ratios.isoc.counts.poiss$Off_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"],
  paired = TRUE
)
```

**Using the standard normalization procedure (i.e. dividing by bin width), we find that there are not more on-isochronous ratios than off-isochronous ratios in eastern chimpanzees.**

#### Western chimpanzees

```{r isoc-counts-wilcoxon-wac-poiss}
wilcox.test(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  ratios.isoc.counts.poiss$Off_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  paired = TRUE
)
```

**Using the standard normalization procedure (i.e. dividing by bin width), we find that there are not more on-isochronous ratios than off-isochronous ratios in western chimpanzees.**

#### Are there more on-isochronous ratios in western than in eastern chimpanzees?

```{r isoc-counts-wilcoxon-eac-vs-wac-poiss}
wilcox.test(
  ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"],
  ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"]
)
print(paste("Median for East:", median(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Eastern"])))
print(paste("Median for West:", median(ratios.isoc.counts.poiss$On_isochronous_norm[ratios.isoc.counts.poiss$Subspecies == "Western"])))
```

**Using the standard normalization procedure (i.e. dividing by bin width), we find that there are more on-isochronous ratios in western chimpanzees than in eastern chimpanzees.**

# References

<div id="refs"></div>